---
title: "HH16_HH31_new_analyses"
output: html_document
date: "2026-05-11"
---

This notebook covers QC metrics, clone composition, and neuronal fate analyses for the injHH16-collHH31 chicken spinal cord LARRY dataset (5 samples: BOE5073A25–A29; injection at HH16, collection at HH31).

**Data required:**

- `preprocessing_larry_chicken_NT_injHH16_collHH31.rds` — preprocessed, annotated Seurat object (from the preprocessing pipeline)

**Outputs:**

- `Table1_clone_QC_HH16HH31.xlsx` — per-sample clone QC summary table
- `Table2_clone_composition_neuronal_fate_HH16HH31.xlsx` — clone class and neuronal fate tables (5 sheets)
- `LARRY_detection_summary_per_sample_HH16HH31.csv` — LARRY detection rates per cell type per sample (**required input for Figure1-S1-S2_new.Rmd**)
- `LARRY_detection_summary_overall_HH16HH31.csv` — LARRY detection rates per cell type overall
- `clone_size_histogram.pdf` — histogram of clone size distribution
- `LARRY_detection_barplot.pdf` — stacked barplot of LARRY barcode detection rate

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r init}
# Load all required packages
library(dplyr)
library(Seurat)
library(readr)
library(tidyverse)
library(scCustomize)
library(ggplot2)
library(ggrepel)

```

## Parameters

```{r parameters}
# Define experiment identifiers and derived file prefix
tissue           <- "NT"
injection_stage  <- "HH16"
collection_stage <- "HH31"

experiment_name  <- paste0(tissue, "_inj", injection_stage, "_coll", collection_stage)
file_prefix      <- paste0("preprocessing_larry_chicken_", experiment_name)

cat("Experiment:", experiment_name, "\n")
```

## Paths to upload Seurat

```{r paths}
# Define input and output directories; create seurat output dir if absent
data_path            <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/data/"
anndata_seurat_path  <- file.path(data_path, "anndata_seurat")
seurat_path          <- file.path(data_path, "seurat")

dir.create(seurat_path, showWarnings = FALSE, recursive = TRUE)
```

## Output paths
```{r output-paths}
# Define plot and data output directories
plots = "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/HH16_HH31_new_May_2026"

data = "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/HH16_HH31_new_May_2026"

```

## Read and prepare Seurat object
```{r load-seurat}
# Load RDS, standardise metadata column names, add experiment/LARRY/clone metadata columns
seur_obj <- read_rds(file.path(seurat_path, paste0(file_prefix, ".rds")))

# Rename columns to match standard naming
colnames(seur_obj@meta.data)[colnames(seur_obj@meta.data) == "cell_type_final"] <- "Cell.type"
colnames(seur_obj@meta.data)[colnames(seur_obj@meta.data) == "clone"]           <- "exp_clone"

# Column 1: "experiment" — all samples are HH31 collection
seur_obj@meta.data$experiment <- case_when(
  seur_obj@meta.data$sample %in% c("BOE5073A25", "BOE5073A26", "BOE5073A27",
                                   "BOE5073A28", "BOE5073A29") ~ "HH31",
  TRUE ~ NA_character_
)

# Column 2: "samples_LARRY"
seur_obj@meta.data$samples_LARRY <- case_when(
  seur_obj@meta.data$sample %in% c("BOE5073A26", "BOE5073A27",
                                   "BOE5073A28", "BOE5073A29") ~ "LARRY",
  seur_obj@meta.data$sample == "BOE5073A25"                    ~ "control",
  TRUE ~ NA_character_
)

# Column 3: "clone_id_detected"
seur_obj@meta.data <- seur_obj@meta.data %>%
  mutate(clone_id_detected = case_when(
    experiment == "HH31" & !is.na(exp_clone) ~ "detected",
    experiment == "HH31" &  is.na(exp_clone) ~ "not_detected",
    TRUE ~ NA_character_
  ))

# Standardise Cell.type — replace NA with Unknown
seur_obj@meta.data$Cell.type <- seur_obj@meta.data$Cell.type %>%
  replace_na("Unknown")

# Clone size
clone_sizes <- seur_obj@meta.data %>%
  filter(!is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

meta_with_clone_status <- seur_obj@meta.data %>%
  left_join(clone_sizes, by = "exp_clone") %>%
  mutate(multi_cell_clone = case_when(
    clone_size > 1  ~ "multi-cell_clone",
    clone_size == 1 ~ "single-cell_clone",
    TRUE            ~ NA_character_
  ))

rownames(meta_with_clone_status) <- rownames(seur_obj@meta.data)

seur_obj <- AddMetaData(seur_obj,
                        metadata = meta_with_clone_status$multi_cell_clone,
                        col.name = "multi_cell_clone")

```

## Clone QC, composition & neuronal fate

### Table 1 - Clone QC summary
```{r clone-qc-table1-hh1631}
# Compute per-sample and overall clone QC stats; save as Table1_clone_QC_HH16HH31.xlsx
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
library(openxlsx)

celltype_order_hh1631 <- c("RP", "dp1", "dp2", "dpL", "dp6", "p0", "p1", "p2", "pMN", "p3", "FP",
                            "OPCs", "dI1", "dI2", "dI3", "dI4", "dI5", "immature_V0_dI6", "dI6",
                            "V0", "V1", "V2a", "V2b", "MN", "V3", "unclear", "Unknown")

# All detected clones (single + multi) per sample
clone_sizes_all <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  group_by(sample, exp_clone) %>%
  summarise(clone_size = n(), .groups = "drop")

# Detection rate per sample
detection_samp <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  group_by(sample) %>%
  summarise(
    total_cells          = n(),
    cells_LARRY_detected = sum(!is.na(exp_clone)),
    pct_detected         = round(100 * cells_LARRY_detected / total_cells, 1),
    .groups = "drop"
  )

# Per-sample clone stats
stats_per_sample <- clone_sizes_all %>%
  group_by(sample) %>%
  summarise(
    n_clones      = n_distinct(exp_clone),
    mean_size     = round(mean(clone_size), 1),
    median_size   = median(clone_size),
    max_size      = max(clone_size),
    n_single_cell = sum(clone_size == 1),
    n_multi_cell  = sum(clone_size > 1),
    pct_single    = round(100 * n_single_cell / n_clones, 1),
    pct_multi     = round(100 * n_multi_cell  / n_clones, 1),
    .groups = "drop"
  ) %>%
  left_join(detection_samp, by = "sample")

# Overall row
overall_clone_stats <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  group_by(exp_clone) %>%
  summarise(clone_size = n(), .groups = "drop") %>%
  summarise(
    n_clones      = n(),
    mean_size     = round(mean(clone_size), 1),
    median_size   = median(clone_size),
    max_size      = max(clone_size),
    n_single_cell = sum(clone_size == 1),
    n_multi_cell  = sum(clone_size > 1),
    pct_single    = round(100 * n_single_cell / n_clones, 1),
    pct_multi     = round(100 * n_multi_cell  / n_clones, 1)
  )

overall_detect <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  summarise(
    total_cells          = n(),
    cells_LARRY_detected = sum(!is.na(exp_clone)),
    pct_detected         = round(100 * cells_LARRY_detected / total_cells, 1)
  )

overall_row <- bind_cols(tibble(sample = "Overall"), overall_clone_stats, overall_detect)
summary_per_sample_full <- bind_rows(stats_per_sample, overall_row)

wb1 <- createWorkbook()
addWorksheet(wb1, "summary_per_sample")
writeData(wb1, "summary_per_sample", summary_per_sample_full)
saveWorkbook(wb1, file.path(data, "Table1_clone_QC_HH16HH31.xlsx"), overwrite = TRUE)
cat("Table1_clone_QC_HH16HH31.xlsx saved\n")
```

### Clone size histogram
```{r clone-size-histogram-hh1631}
# Plot histogram of clone sizes with median and mean annotations; save as PDF
clone_sizes_hist <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

annot <- clone_sizes_hist %>%
  summarise(med = median(clone_size), mean = round(mean(clone_size), 2), n = n())

ggplot(clone_sizes_hist, aes(x = clone_size)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = annot$med, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("n = ", annot$n, "\nmedian = ", annot$med, "\nmean = ", annot$mean),
           hjust = 1.1, vjust = 1.2, size = 3.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  theme_minimal(base_size = 13) +
  labs(title = "Clone size distribution (all LARRY clones — HH31)",
       x = "Clone size (number of cells)", y = "Number of clones")

ggsave(filename = "clone_size_histogram.pdf", plot = last_plot(), path = plots,
       width = 18, height = 12, units = "cm", dpi = 300, bg = "transparent")
```

### Table 2 - Clone composition & neuronal fate
```{r clone-table2-hh1631}
# Classify multi-cell clones by lineage composition and neuronal fate; save as Table2 Excel
library(openxlsx)

progenitors_hh1631 <- c("RP", "dp1", "dp2", "dpL", "dp6", "p0", "p1", "p2", "pMN", "p3", "FP")
neurons_hh1631     <- c("dI1", "dI2", "dI3", "dI4", "dI5", "dI6", "immature_V0_dI6",
                         "V0", "V1", "V2a", "V2b", "MN", "V3")
glia_hh1631        <- c("OPCs")
# unclear + Unknown → excluded_only

# Multi-cell clones only
clone_cells_multi <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", multi_cell_clone == "multi-cell_clone") %>%
  select(exp_clone, Cell.type)

# Clone class per clone
clone_class_df <- clone_cells_multi %>%
  group_by(exp_clone) %>%
  summarise(
    has_prog = any(Cell.type %in% progenitors_hh1631),
    has_neur = any(Cell.type %in% neurons_hh1631),
    has_glia = any(Cell.type %in% glia_hh1631),
    .groups = "drop"
  ) %>%
  mutate(clone_class = case_when(
    has_prog & !has_neur & !has_glia ~ "progenitor_only",
    !has_prog & has_neur & !has_glia ~ "neuron_only",
    !has_prog & !has_neur & has_glia ~ "glia_only",
    has_prog & has_neur & !has_glia  ~ "progenitor_neuron",
    has_prog & !has_neur & has_glia  ~ "glia_progenitor",
    !has_prog & has_neur & has_glia  ~ "glia_neuron",
    has_prog & has_neur & has_glia   ~ "mixed_all",
    TRUE                             ~ "excluded_only"
  ))

class_levels <- c("progenitor_only", "neuron_only", "glia_only",
                  "progenitor_neuron", "glia_progenitor", "glia_neuron",
                  "mixed_all", "excluded_only")

clone_class_overall <- clone_class_df %>%
  count(clone_class) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(factor(clone_class, levels = class_levels))

# Neuronal fate — V2 collapse (V2a + V2b → V2, same as Fig 1 chicken)
neuron_types_v2 <- c("dI1", "dI2", "dI3", "dI4", "dI5", "dI6", "immature_V0_dI6",
                      "V0", "V1", "V2", "MN", "V3")

clone_fate_v2 <- clone_cells_multi %>%
  mutate(Cell.type_v2 = case_when(
    Cell.type %in% c("V2a", "V2b") ~ "V2",
    TRUE ~ Cell.type
  )) %>%
  filter(Cell.type_v2 %in% neuron_types_v2) %>%
  group_by(exp_clone) %>%
  summarise(
    neuron_types   = list(sort(unique(Cell.type_v2))),
    n_neuron_types = n_distinct(Cell.type_v2),
    .groups = "drop"
  ) %>%
  mutate(neuron_fate = case_when(
    n_neuron_types == 1 ~ "single_neuronal_fate",
    n_neuron_types > 1  ~ "multi_neuronal_fate"
  ))

# fate_overall: neuronal clones only as denominator
fate_overall <- clone_fate_v2 %>%
  count(neuron_fate) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(factor(neuron_fate, levels = c("single_neuronal_fate", "multi_neuronal_fate")))

# fate_all_clones: all multicell clones as denominator (reviewer question)
fate_all_clones_overall <- clone_class_df %>%
  left_join(clone_fate_v2 %>% select(exp_clone, neuron_fate), by = "exp_clone") %>%
  mutate(fate_category = case_when(
    is.na(neuron_fate) ~ "no_neuronal_fate",
    TRUE               ~ neuron_fate
  )) %>%
  count(fate_category) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(factor(fate_category, levels = c("no_neuronal_fate",
                                           "single_neuronal_fate",
                                           "multi_neuronal_fate")))

# fate_by_type: per neuron type, how many clones contain it
fate_by_type <- clone_cells_multi %>%
  mutate(Cell.type_v2 = case_when(
    Cell.type %in% c("V2a", "V2b") ~ "V2",
    TRUE ~ Cell.type
  )) %>%
  filter(Cell.type_v2 %in% neuron_types_v2) %>%
  group_by(exp_clone, Cell.type_v2) %>%
  summarise(n_cells = n(), .groups = "drop") %>%
  group_by(Cell.type_v2) %>%
  summarise(
    n_clones_with_type = n_distinct(exp_clone),
    .groups = "drop"
  ) %>%
  mutate(pct_of_neuronal_clones = round(100 * n_clones_with_type / nrow(clone_fate_v2), 1)) %>%
  arrange(factor(Cell.type_v2, levels = neuron_types_v2))

# Multi-fate combos
multi_fate_combos <- clone_fate_v2 %>%
  filter(neuron_fate == "multi_neuronal_fate") %>%
  mutate(combo = sapply(neuron_types, paste, collapse = "+")) %>%
  count(combo, sort = TRUE) %>%
  mutate(pct = round(100 * n / sum(n), 1))

wb2 <- createWorkbook()
addWorksheet(wb2, "clone_class_overall")
writeData(wb2, "clone_class_overall", clone_class_overall)
addWorksheet(wb2, "fate_overall")
writeData(wb2, "fate_overall", fate_overall)
addWorksheet(wb2, "fate_all_clones_overall")
writeData(wb2, "fate_all_clones_overall", fate_all_clones_overall)
addWorksheet(wb2, "fate_by_type")
writeData(wb2, "fate_by_type", fate_by_type)
addWorksheet(wb2, "multi_fate_combos")
writeData(wb2, "multi_fate_combos", multi_fate_combos)
saveWorkbook(wb2, file.path(data, "Table2_clone_composition_neuronal_fate_HH16HH31.xlsx"), overwrite = TRUE)
cat("Table2_clone_composition_neuronal_fate_HH16HH31.xlsx saved\n")
```

## LARRY detection rate per cell type
```{r dropout-hh1631}
# Compute LARRY detection rates per cell type (per sample and overall); save as CSVs
# Per cell type per sample (double-filter NA pattern)
dropout_per_sample <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(Cell.type)) %>%
  mutate(Cell.type = factor(Cell.type, levels = celltype_order_hh1631)) %>%
  filter(!is.na(Cell.type)) %>%
  group_by(sample, Cell.type) %>%
  summarise(
    total_cells          = n(),
    cells_LARRY_detected = sum(!is.na(exp_clone)),
    pct_detected         = round(100 * cells_LARRY_detected / total_cells, 1),
    .groups = "drop"
  )

# Per cell type overall
dropout_overall <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(Cell.type)) %>%
  mutate(Cell.type = factor(Cell.type, levels = celltype_order_hh1631)) %>%
  filter(!is.na(Cell.type)) %>%
  group_by(Cell.type) %>%
  summarise(
    total_cells          = n(),
    cells_LARRY_detected = sum(!is.na(exp_clone)),
    pct_detected         = round(100 * cells_LARRY_detected / total_cells, 1),
    .groups = "drop"
  )

write.csv(dropout_per_sample, file.path(data, "LARRY_detection_summary_per_sample_HH16HH31.csv"), row.names = FALSE)
write.csv(dropout_overall,    file.path(data, "LARRY_detection_summary_overall_HH16HH31.csv"),    row.names = FALSE)
cat("Detection summary CSVs saved\n")
```

## LARRY detection barplot
```{r detection-barplot-hh1631}
# Stacked detected / not_detected, one bar for the whole dataset
detection_plot_df <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  mutate(clone_id_detected = ifelse(!is.na(exp_clone), "detected", "not_detected")) %>%
  group_by(clone_id_detected) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    percent_cells     = round(100 * n / sum(n), 1),
    dataset           = "HH16 → HH31",
    clone_id_detected = factor(clone_id_detected, levels = c("not_detected", "detected"))
  ) %>%
  arrange(clone_id_detected) %>%                      # not_detected first → bottom
  mutate(y_label = cumsum(percent_cells) - percent_cells / 2)

ggplot(detection_plot_df, aes(x = dataset, y = percent_cells, fill = clone_id_detected)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    aes(label = round(percent_cells, digits = 1), y = y_label),
    colour = "black", size = 8, hjust = 0.5, vjust = 0.5
  ) +
  scale_fill_manual(
    values = c("not_detected" = "grey90", "detected" = "darkcyan"),
    labels = c("not_detected" = "Not detected", "detected" = "Detected")
  ) +
  labs(title = "LARRY barcode detection") +
  ylab("% of cells") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    plot.title   = element_text(face = "bold", size = 24, vjust = 0.5, hjust = 0.5),
    axis.title.y = element_text(face = "bold", size = 18),
    axis.text    = element_text(size = 14)
  )

ggsave(filename = "LARRY_detection_barplot.pdf", plot = last_plot(), path = plots,
       width = 3, height = 12, units = "cm", dpi = 300, bg = "transparent")
```
