---
title: "injHH7-collHH18 LARRY NT — Cluster Inspection & Annotation"
author: "Giulia"
date: "`r Sys.Date()`"
output: html_document
---

This notebook covers cluster inspection and cell type annotation for the injHH7-collHH18 chicken spinal cord LARRY dataset (injHH7-collHH18 dataset; injection at HH7-9, collection at HH18).

**Data required:**

- `LARRY_NT_integrated_07-09-2025_annotated.rds` — integrated Seurat object with existing annotation columns (primary input)
- `LARRY_integrated_switched_somites_07-09-2025.rds` — full (pre-subset) RDS used to transfer annotation columns (used in the one-time filtering step)

**Outputs:**

- `LARRY_NT_integrated_07-09-2025_filtered_annotated.rds` — NT-only Seurat object after removing non-NT germ layers and partially-retained clusters
- `LARRY_NT_integrated_07-09-2025_annotated.rds` — updated RDS with `cell_type_fine`, `cell_type_coarse`, and `cell_type_collapsed` columns written back
- `LARRY_NT_integrated_07-09-2025_annotated_metadata.csv` — flat metadata export (clone, somites, annotation columns)
- `Table1_clone_QC_injHH7-collHH18.xlsx` — per-sample clone QC summary
- `Table2_clone_domain_diversity_injHH7-collHH18.xlsx` — progenitor domain fate and multi-domain combo tables
- `LARRY_detection_summary_per_sample_injHH7-collHH18.csv` — LARRY detection rates per cell type per sample
- `LARRY_detection_summary_overall_injHH7-collHH18.csv` — LARRY detection rates per cell type overall
- `clone_size_histogram.pdf` — clone size distribution plot
- `LARRY_detection_barplot.pdf` — stacked barplot of LARRY barcode detection rate
- `UMAP-RNA_snn_res25-dims13.png` — UMAP coloured by clustering resolution 25
- `Dotplot-res10.png` — DotPlot of marker genes at resolution 10
- Individual FeaturePlot PNGs per marker gene (saved to plots directory)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Libraries
```{r libs}
# Load all required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(tidyverse)
library(scCustomize)
library(ggplot2)
library(ggrepel)

```

## Paths
```{r paths}
# Define input data and output plot directories
data  <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/injHH7-collHH18_LARRY/data/"
plots <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/injHH7-collHH18_LARRY/plots/"
```

## Load RDS
```{r load}
# Load the annotated integrated Seurat object and print metadata overview
#seur_obj <- read_rds(file.path(data, "LARRY_NT_integrated_07-09-2025.rds"))

seur_obj <- read_rds(file.path(data, "LARRY_NT_integrated_07-09-2025_annotated.rds"))

DefaultAssay(seur_obj) <- "RNA"

cat("=== Metadata columns ===\n"); print(colnames(seur_obj@meta.data))
cat("=== Reductions ===\n");       print(names(seur_obj@reductions))
cat("=== n cells ===\n");          print(ncol(seur_obj))
```

## Metadata setup
```{r metadata}
# Standardise metadata: rename clone column, derive experiment/samples_LARRY/clone_id_detected/multi_cell_clone
# Rename clone → exp_clone if needed (fresh RDS has "clone")
if ("clone" %in% colnames(seur_obj@meta.data)) {
  seur_obj@meta.data$exp_clone <- seur_obj@meta.data$clone
  seur_obj@meta.data$clone     <- NULL
}

# experiment: use somites column which already has correct labels
seur_obj@meta.data$experiment <- seur_obj@meta.data$somites

# samples_LARRY
seur_obj@meta.data$samples_LARRY <- case_when(
  seur_obj@meta.data$sample %in% c("LIB6512A4", "LIB6512A14") ~ "control",
  seur_obj@meta.data$sample %in% c("LIB6512A1", "LIB6512A2", "LIB6512A3",
                                   "LIB6512A10", "LIB6512A11", "LIB6512A12",
                                   "LIB6512A13") ~ "LARRY",
  TRUE ~ NA_character_
)

# clone_id_detected
seur_obj$clone_id_detected <- case_when(
  seur_obj$samples_LARRY == "control" ~ "control",
  !is.na(seur_obj$exp_clone)          ~ "detected",
  is.na(seur_obj$exp_clone)           ~ "not_detected",
  TRUE                                ~ NA_character_
)

# multi_cell_clone (safe: named vector, no left_join)
clone_size_lookup <- seur_obj@meta.data %>%
  filter(!is.na(exp_clone)) %>%
  count(exp_clone, name = "n") %>%
  tibble::deframe()

seur_obj$multi_cell_clone <- case_when(
  is.na(seur_obj$exp_clone)                 ~ NA_character_,
  clone_size_lookup[seur_obj$exp_clone] > 1 ~ "multi-cell_clone",
  TRUE                                      ~ "single-cell_clone"
)

table(seur_obj$experiment, useNA = "ifany")
table(seur_obj$samples_LARRY, useNA = "ifany")
table(seur_obj$multi_cell_clone, useNA = "ifany")
```

## One-time NT filtering and annotation transfer (run once then skip and reload the full rds)
```{r filter-transfer-save}
# Remove non-NT cell types, transfer annotation from full RDS, remove partially-retained clusters, save filtered object
# ── 1. Remove non-NT cell types ──────────────────────────────────────────────
table(seur_obj$germ.layers, useNA = "ifany")   # inspect first

to_remove  <- c("brain", "brain neuron", "optic cup", "hindbrain", "tailbud", "preneural")
cells_keep <- colnames(seur_obj)[!seur_obj$germ.layers %in% to_remove]
seur_obj   <- subset(seur_obj, cells = cells_keep)

cat("Cells remaining:", ncol(seur_obj), "\n")


# ── 2. Transfer annotation column(s) from the full RDS ───────────────────────
seur_full <- readRDS("/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/injHH7-collHH18_LARRY/data/LARRY_integrated_switched_somites_07-09-2025.rds")

# check what annotation columns are available
colnames(seur_full@meta.data)

cell_type_cols <- grep("^cell.type", colnames(seur_full@meta.data), value = TRUE)

for (col in cell_type_cols) {
  cat("\n===", col, "===\n")
  print(sort(unique(seur_full@meta.data[[col]])))
}

# transfer — works because seur_obj barcodes are a subset of seur_full barcodes
shared_cells <- intersect(colnames(seur_obj), colnames(seur_full))
cat("Barcodes matched:", length(shared_cells), "of", ncol(seur_obj), "\n")

# replace "Cell.type" with whichever column(s) you want
seur_obj$cell.type1 <- seur_full@meta.data[colnames(seur_obj), "cell.type1"]


#Check partial removed clusrters and remove them
orig_sizes    <- table(seur_full@meta.data[["RNA_snn_res.10"]])
current_sizes <- table(seur_obj@meta.data[["RNA_snn_res.10"]])

retention <- data.frame(
     cluster      = names(orig_sizes),
     orig         = as.numeric(orig_sizes),
     current      = as.numeric(current_sizes[names(orig_sizes)])
 ) %>%
     mutate(current       = ifelse(is.na(current), 0, current),
            pct_retained  = round(100 * current / orig, 1)) %>%
     arrange(pct_retained)

print(retention)

remove_clusters <- c("26", "29", "30", "46", "75")
cells_keep <- colnames(seur_obj)[!seur_obj@meta.data[["RNA_snn_res.10"]] %in% remove_clusters]
seur_obj   <- subset(seur_obj, cells = cells_keep)
cat("Cells remaining:", ncol(seur_obj), "\n")

rm(seur_full)

# ── 3. Save ───────────────────────────────────────────────────────────────────
saveRDS(seur_obj, file.path(data, "LARRY_NT_integrated_07-09-2025_filtered_annotated.rds"))

```

---

## Cluster inspection

### Overview at different resolutions
```{r dimplot-coarse, fig.width=20, fig.height=8}
# Plot UMAPs at resolutions 20, 25, and 35 to compare cluster granularity
p1 <- DimPlot(seur_obj, group.by = "RNA_snn_res.20", reduction = "umap", dims = c(1,3),
              label = TRUE, label.size = 3, pt.size = 0.6) +
      ggtitle("res 20") + NoLegend()

p2 <- DimPlot(seur_obj, group.by = "RNA_snn_res.25", reduction = "umap", dims = c(1,3),
              label = TRUE, label.size = 3, pt.size = 0.6) +
      ggtitle("res 25") + NoLegend()

p3 <- DimPlot(seur_obj, group.by = "RNA_snn_res.35", reduction = "umap", dims = c(1,3),
              label = TRUE, label.size = 3, pt.size = 0.6) +
      ggtitle("res 35") + NoLegend()

p2

ggsave(filename = "UMAP-RNA_snn_res25-dims13.png", plot = p2, path = plots,
       scale = 1, width = 35, height = 26, units = "cm",
       dpi = 300, bg = "transparent")

```

### Save UMAP DimPlots
```{r save-umap-dimplot}
# Generate and save a labelled UMAP coloured by resolution 25 clusters
library(scCustomize)

n_clusters <- length(unique(seur_obj@meta.data[["RNA_snn_res.25"]]))
cols <- DiscretePalette_scCustomize(num_colors = n_clusters, palette = "varibow")

p <- DimPlot(seur_obj, group.by = "RNA_snn_res.25", reduction = "umap", dims = c(1, 3),
             label = TRUE, label.size = 8, label.box = TRUE, repel = TRUE, pt.size = 1.5,
             cols = cols) +
     ggtitle("RNA_snn_res.25") +
     guides(color = guide_legend(ncol = 2, override.aes = list(size = 4)))

ggsave(filename = "UMAP-RNA_snn_res25-dims13.png", plot = p, path = plots,
       scale = 1, width = 35, height = 26, units = "cm",
       dpi = 300, bg = "transparent")

```

---

## FeaturePlots — Progenitor markers
```{r fp-progenitors, fig.width=15, fig.height=12}
# FeaturePlots for dorsal and ventral progenitor marker genes
progenitors <- c("SOX2", "SHH", "FOXA2","ARX", "NKX2-2", "NKX6-2", "OLIG2",
                 "PAX6", "IRX3", "DBX2", "PAX3", "PAX7", "LMX1A", "WNT1",
                 "SP8", "PRDM12", "ASCL1", "MSX1"
                 # ,"PDGFRA", "SOX10", "OTX1", "EN2"
                 )

progenitors <- c("SHH", "FOXA2","ARX",  "LMX1A", "WNT1", "MSX1", "OTX1", "EN2"
                 )

FeaturePlot(seur_obj,
            dims = c(2, 3),
            features = intersect(progenitors, rownames(seur_obj)),
            reduction = "umap", ncol = 2,
            min.cutoff = "q10", max.cutoff = "q95", order = TRUE)
```

## FeaturePlots — Neuron markers
```{r fp-neurons, fig.width=15, fig.height=20}
# FeaturePlots for interneuron and motor neuron marker genes
neurons <- c("TUBB3", "ISL1", "ISL2", "LHX3", "MNX1", "FOXP1",
             "EN1", "EVX1", "GATA2", "GATA3", "VSX2", "LMX1B",
             "LHX1", "LHX5", "PAX2", "TLX3", "LBX1", "BARHL2",
             "SLC17A6", "SLC32A1", "GAD1", "SLC6A5",
             "ONECUT1", "ONECUT2", "NFIA", "NFIB", "ZFHX3")

FeaturePlot(seur_obj, features = intersect(neurons, rownames(seur_obj)),
            reduction = "umap", ncol = 5,
            min.cutoff = "q10", max.cutoff = "q99", order = TRUE)
```

## FeaturePlots — save individually
```{r fp-save-loop}
# Loop over a curated marker list and save individual FeaturePlot PNGs to the plots directory
# Change this list to whatever you want to plot
markers_to_save <- c(
  "SOX2", "LMX1A", "MSX1", "WNT1",  "PAX3",
  "OLIG3",  "PAX7", "IRX3","PAX6",
  "GSX2", "ASCL1", "GBX1", "PTF1A", "DBX2",
  "SP8", "PRDM12", "FOXN4", "NKX6-2", "OLIG2", "NKX2-2",  "NKX2-8", "FOXA2",
  "ARX", "SHH", "FERD3L",

  "POU4F1", "LHX2", "LHX9", "BARHL1",
  "LHX1", "LHX5", "ISL1", "TLX3", "LBX1",
  "PAX2", "GBX1", "LMX1B",
  "DMRT3", "WT1", "IRX1", "IRX2", "EVX1",  "EN1", "LHX3",
  "SOX14", "SOX21","VSX2", "SHOX2",  "TAL1", "GATA2", "GATA3",
   "MNX1",  "ISL2", "SLC10A4", "SIM1")



for (gene in markers_to_save) {
  tryCatch({
    p <- FeaturePlot(seur_obj,
                     reduction = "umap",
                     dims = c(2, 3),
                     features = gene,
                     pt.size = 0.7,
                     min.cutoff = "q10", max.cutoff = "q95",
                     order = TRUE) +
         scale_x_discrete(breaks = NULL) +
         scale_y_discrete(breaks = NULL) +
         labs(title = gene)
    ggsave(filename = paste0(gene, "_Featureplot-dims2-3.png"), plot = p, path = plots,
           scale = 1, width = 17, height = 16, units = "cm",
           dpi = 300, bg = "transparent")
  }, error = function(e) message(paste("Skipping", gene, ":", e$message)))
}
cat("Done saving featureplots.\n")
```

## Check clusters removed by subsetting
```{r check-missing-clusters}
# Print cluster IDs that are absent from the current object after subsetting
all_levels <- levels(seur_obj@meta.data[["RNA_snn_res.10"]])
present    <- as.character(unique(seur_obj@meta.data[["RNA_snn_res.10"]]))
missing    <- setdiff(all_levels, present)
cat("Missing clusters:", paste(sort(as.numeric(missing)), collapse = ", "), "\n")

```

## DotPlot for cluster inspection
```{r dotplot-inspect, fig.width=18, fig.height=14}
# DotPlot of curated marker genes across all clusters at the chosen resolution
# Change resolution column here once you've picked one
chosen_res <- "RNA_snn_res.10"
Idents(seur_obj) <- chosen_res

genes_dotplot <- c(
  "SOX2", "LMX1A", "MSX1", "MSX2", "WNT1",  "PAX3",
  "OLIG3",  "PAX7", "IRX3","PAX6",
  "GSX2", "ASCL1", "GBX1", "GBX2", "PTF1A", "DBX2",
  "SP8", "PRDM12", "FOXN4", "NKX6-2", "OLIG2", "NKX2-2",  "NKX2-8", "FOXA2",
  "ARX", "SHH", "FERD3L",

  "POU4F1", "LHX2", "LHX9", "BARHL1",
  "LHX1", "LHX5", "ISL1", "TLX3", "LBX1",
  "PAX2", "GBX1", "LMX1B",
  "DMRT3", "WT1", "IRX1", "IRX2", "EVX1",  "EN1", "LHX3",
  "SOX14", "SOX21","VSX2", "SHOX2",  "TAL1", "GATA2", "GATA3",
   "MNX1",  "ISL2", "SLC10A4", "SIM1")

DotPlot(seur_obj,
        features = intersect(genes_dotplot, rownames(seur_obj)),
        dot.scale = 5) +
  scale_color_viridis_c(option = "cividis") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        axis.text.y = element_text(size = 8)) +
  labs(x = NULL, y = "Cluster", size = "% Expressing", color = "Avg Expr")


ggsave(filename = "Dotplot-res10.png", plot = last_plot(), path = plots,
       scale = 1, width = 35, height = 22, units = "cm",
       dpi = 300, bg = "transparent")

```

## Check specific clusters on UMAP
```{r highlight-clusters}
# Highlight selected clusters on UMAP to inspect their spatial distribution
clusters_to_color <- c("14", "27", "31", "39", "40", "81")   # replace with your clusters

cells_list <- lapply(clusters_to_color, function(cl) {
  colnames(seur_obj)[seur_obj@meta.data[["RNA_snn_res.10"]] == cl]
})
names(cells_list) <- clusters_to_color

DimPlot(seur_obj, reduction = "umap", dims = c(2, 3),
        cells.highlight = cells_list,
        cols.highlight = DiscretePalette_scCustomize(num_colors = length(clusters_to_color), palette = "varibow"),
        cols = "lightgrey",
        pt.size = 1.5,
        sizes.highlight = 2) +
  ggtitle("Highlighted clusters")


ggsave(filename = "UMAP-clusters14-27-31-39-40-81-dim23.png", plot = last_plot(), path = plots,
       scale = 1, width = 35, height = 26, units = "cm",
       dpi = 300, bg = "transparent")

```

## Annotation
```{r assign-annotation}
# Apply cluster-to-cell-type lookup table and assign cell_type_fine and cell_type_coarse columns
# ── Annotation lookup ─────────────────────────────────────────────────────────
annotation <- tibble::tribble(
  ~cluster, ~cell_type_fine,  ~cell_type_coarse,
  "0",      "dp6/dp5",        "dp6",
  "2",      "p0",             "p0",
  "4",      "pMN/p3",         "pMN",
  "5",      "p2/p1",          "p0/p1",
  "7",      "p0",             "p0/p1",
  "8",      "p1",             "p0/p1",
  "10",     "pMN/p2",         "pMN/p2",
  "11",     "pMN/p2",         "pMN/p2",
  "12",     "p0",             "p0/p1",
  "13",     "dp1",            "dp1/2",
  "14",     "dp4/5",          "dp3/5",
  "17",     "dp6",            "dp6",
  "18",     "p1/p2",          "p0/p1",
  "19",     "dp1",            "dp1/2",
  "20",     "pMN",            "pMN/p2",
  "21",     "dp1",            "dp1/2",
  "25",     "p2",             "pMN/p2",
  "27",     "dp3/4",          "dp3/5",
  "31",     "dp3/4",          "dp3/5",
  "32",     "V0/V1",          "p0/p1",
  "33",     "p1",             "p0/p1",
  "34",     "pMN",            "pMN/p2",
  "35",     "dp4/5",          "dp3/5",
  "36",     "pMN",            "pMN/p2",
  "38",     "MN",             "pMN/p2",
  "39",     "dp3",            "dp3/5",
  "40",     "dp3/dp4",        "dp3/5",
  "41",     "p2",             "pMN/p2",
  "42",     "p0",             "p0/p1",
  "43",     "MN",             "pMN/p2",
  "44",     "p0/p1",          "p0/p1",
  "45",     "dp3/4",          "dp3/5",
  "49",     "MN",             "pMN/p2",
  "51",     "p0/dp6",         "p0/p1",
  "53",     "dp4/5",          "dp3/5",
  "54",     "p0/p1",          "p0/p1",
  "55",     "dI4/6",          "dp3/5",
  "57",     "dI5",            "dp3/5",
  "59",     "dp6",            "dp6",
  "60",     "V3",             "p3",
  "62",     "dp1",            "dp1/2",
  "63",     "dp6",            "dp6",
  "64",     "p0",             "p0/p1",
  "65",     "RP",             "RP",
  "66",     "p0",             "p0/p1",
  "67",     "dp2",            "dp1/2",
  "69",     "FP",             "FP",
  "70",     "p1",             "p0/p1",
  "71",     "p3",             "p3",
  "72",     "p1",             "p0/p1",
  "73",     "dp5",            "dp3/5",
  "74",     "p0/p1",          "p0/p1",
  "76",     "dI1",            "dp1/2",
  "78",     "Unclear",        "Unclear",
  "79",     "V2",             "pMN/p2",
  "80",     "V3",             "p3",
  "81",     "dp4/5",          "dp3/5",
  "82",     "MN",             "pMN/p2",
  "85",     "p0",             "p0/p1",
  "87",     "p0",             "p0/p1",
  "88",     "Unclear",        "Unclear"
)

lookup_fine   <- setNames(annotation$cell_type_fine,   annotation$cluster)
lookup_coarse <- setNames(annotation$cell_type_coarse, annotation$cluster)

res10 <- as.character(seur_obj@meta.data[["RNA_snn_res.10"]])
res25 <- as.character(seur_obj@meta.data[["RNA_snn_res.25"]])

# ── Assign annotations ────────────────────────────────────────────────────────
seur_obj$cell_type_fine <- unname(case_when(
  res10 == "22" & res25 %in% c("97", "142") ~ "dI2",
  res10 == "22" & res25 == "120"             ~ "dp1",
  res10 == "23" & res25 == "45"              ~ "MN",
  res10 == "23" & res25 %in% c("102", "152") ~ "Unclear",
  TRUE ~ lookup_fine[res10]
))

seur_obj$cell_type_coarse <- unname(case_when(
  res10 == "22" & res25 %in% c("97", "142") ~ "dp1/2",
  res10 == "22" & res25 == "120"             ~ "dp1/2",
  res10 == "23" & res25 == "45"              ~ "pMN/p2",
  res10 == "23" & res25 %in% c("102", "152") ~ "Unclear",
  TRUE ~ lookup_coarse[res10]
))

table(seur_obj$cell_type_fine,   useNA = "ifany")
table(seur_obj$cell_type_coarse, useNA = "ifany")

seur_obj$cell_type_coarse[seur_obj$cell_type_coarse == "pMN"] <- "pMN/p2"
seur_obj$cell_type_coarse[seur_obj$cell_type_coarse == "p0"] <- "p0/p1"
table(seur_obj$cell_type_coarse)


seur_obj$cell_type_fine[seur_obj$cell_type_fine == "dp3/dp4"] <- "dp3/4"
seur_obj$cell_type_fine[seur_obj$cell_type_fine == "p2/p1"] <- "p1/p2"
table(seur_obj$cell_type_fine,   useNA = "ifany")

```

## Collapse cell type labels
```{r collapse-celltypes}
# Recode cell_type_fine into a simplified cell_type_collapsed column for downstream analyses
seur_obj$cell_type_collapsed <- dplyr::recode(
  seur_obj$cell_type_fine,
  "dp3/4"   = "dp4/5",
  "dp5"     = "dp4/5",
  "pMN/p2"  = "pMN",
  "pMN/p3"  = "pMN",
  "p0/p1"   = "p0",
  "dp6/dp5" = "dp6",
  "p0/dp6"  = "dp6",
  "p1/p2"   = "p2",
  "dI4/6"   = "dI4/dI5",
  "dI5"     = "dI4/dI5"
)

table(seur_obj$cell_type_collapsed)
```

## Save annotated RDS
```{r save-annotated-rds}
# Save the fully annotated Seurat object back to disk
saveRDS(seur_obj, file.path(data, "LARRY_NT_integrated_07-09-2025_annotated.rds"))
cat("Saved.\n")
```

---

### Check annotation on UMAP
```{r check-umap-clusters}
# Plot UMAP coloured by resolution 25 clusters to verify annotation coverage
library(scCustomize)

n_clusters <- length(unique(seur_obj@meta.data[["RNA_snn_res.25"]]))
cols <- DiscretePalette_scCustomize(num_colors = n_clusters, palette = "varibow")

p <- DimPlot(seur_obj, group.by = "RNA_snn_res.25", reduction = "umap", dims = c(1, 3),
             label = TRUE, label.size = 8, label.box = TRUE, repel = TRUE, pt.size = 1.5,
             cols = cols) +
     ggtitle("RNA_snn_res.25") +
     guides(color = guide_legend(ncol = 2, override.aes = list(size = 4)))

ggsave(filename = "UMAP-RNA_snn_res25-dims13.png", plot = p, path = plots,
       scale = 1, width = 35, height = 26, units = "cm",
       dpi = 300, bg = "transparent")

```

```{r check-annotation, fig.width=10, fig.height=8}
# Plot UMAP coloured by collapsed cell type labels to confirm annotation spread
n_clusters <- length(unique(seur_obj@meta.data[["cell_type_collapsed"]]))
cols <- DiscretePalette_scCustomize(num_colors = n_clusters, palette = "varibow")

DimPlot(seur_obj, group.by = "cell_type_collapsed", reduction = "umap", dims = c(1, 3),
             label = TRUE, label.size = 8, label.box = TRUE, repel = TRUE, pt.size = 1.5,
             cols = cols) +
     ggtitle("cell_type_collapsed") +
     guides(color = guide_legend(ncol = 2, override.aes = list(size = 4)))


```

## Export annotated metadata CSV
```{r export-metadata-csv}
# Write flat metadata table with clone, somites, and all annotation columns to CSV
seur_obj@meta.data %>%
  tibble::rownames_to_column("index") %>%
  select(index, clone = exp_clone, somites, cell_type_fine, cell_type_coarse, cell_type_collapsed) %>%
  write.csv(file.path(data, "LARRY_NT_integrated_07-09-2025_annotated_metadata.csv"), row.names = FALSE)
```

---

## Clone QC, composition & domain diversity

### Clone QC — Table 1
```{r clone-qc-table}
# Compute per-sample and overall clone QC stats; save as Table1_clone_QC_injHH7-collHH18.xlsx
library(dplyr)
library(tidyr)
library(readr)

if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
library(openxlsx)

data  <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/injHH7-collHH18_LARRY/data/"
plots <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/injHH7-collHH18_LARRY/plots/"

# ── Per-sample clone stats ────────────────────────────────────────────────────
clone_sizes_raw <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, sample, name = "clone_size")

detection_per_sample <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  group_by(sample) %>%
  summarise(
    total_cells = n(),
    cells_LARRY_detected    = sum(clone_id_detected == "detected", na.rm = TRUE),
    pct_detected      = round(100 * cells_LARRY_detected / total_cells, 1),
    .groups = "drop"
  )

clones_per_sample <- clone_sizes_raw %>%
  group_by(sample) %>%
  summarise(
    n_clones      = n_distinct(exp_clone),
    mean_size     = round(mean(clone_size), 2),
    median_size   = median(clone_size),
    max_size      = max(clone_size),
    n_single_cell = sum(clone_size == 1),
    n_multi_cell  = sum(clone_size > 1),
    pct_single    = round(100 * sum(clone_size == 1) / n(), 1),
    pct_multi     = round(100 * sum(clone_size > 1)  / n(), 1),
    .groups = "drop"
  ) %>%
  left_join(detection_per_sample, by = "sample")

# ── Overall row (computed from raw data, not averaged across samples) ─────────
clone_sizes_all <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

larry_cells_all <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY")

table1_overall <- tibble(
  sample            = "Overall",
  n_clones          = nrow(clone_sizes_all),
  mean_size         = round(mean(clone_sizes_all$clone_size), 2),
  median_size       = median(clone_sizes_all$clone_size),
  max_size          = max(clone_sizes_all$clone_size),
  n_single_cell     = sum(clone_sizes_all$clone_size == 1),
  n_multi_cell      = sum(clone_sizes_all$clone_size > 1),
  pct_single        = round(100 * sum(clone_sizes_all$clone_size == 1) / nrow(clone_sizes_all), 1),
  pct_multi         = round(100 * sum(clone_sizes_all$clone_size > 1)  / nrow(clone_sizes_all), 1),
  total_cells = nrow(larry_cells_all),
  cells_LARRY_detected    = sum(larry_cells_all$clone_id_detected == "detected", na.rm = TRUE),
  pct_detected      = round(100 * sum(larry_cells_all$clone_id_detected == "detected", na.rm = TRUE) /
                              nrow(larry_cells_all), 1)
)

table1 <- bind_rows(clones_per_sample, table1_overall)
print(table1)

wb1 <- createWorkbook()
addWorksheet(wb1, "summary_per_sample")
writeData(wb1, "summary_per_sample", table1)
saveWorkbook(wb1, file.path(data, "Table1_clone_QC_injHH7-collHH18.xlsx"), overwrite = TRUE)
cat("Table 1 saved!\n")
```

### Clone size histogram
```{r clone-size-histogram}
# Plot histogram of clone sizes with median and mean annotations; save as PDF
clone_sizes_hist <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

annot <- clone_sizes_hist %>%
  summarise(med = median(clone_size), mean = round(mean(clone_size), 2), n = n())

ggplot(clone_sizes_hist, aes(x = clone_size)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = annot$med, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("n = ", annot$n, "\nmedian = ", annot$med, "\nmean = ", annot$mean),
           hjust = 1.1, vjust = 1.2, size = 3.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  theme_minimal(base_size = 13) +
  labs(title = "Clone size distribution (all LARRY clones — injHH7-collHH18 NT)",
       x = "Clone size (number of cells)", y = "Number of clones")

ggsave(filename = "clone_size_histogram.pdf", plot = last_plot(), path = plots,
       scale = 1, width = 20, height = 15, units = "cm", dpi = 300, bg = "transparent")
```

### Clone domain diversity — Table 2
```{r clone-domain-diversity}
# Classify multi-cell clones by progenitor domain diversity; save as Table2 Excel
# ── Cell type categories (cell_type_collapsed) ────────────────────────────────
# Progenitor domains: all p-domain identities
# Neurons: V0/V1 included as neuronal (V0, V1 are spinal interneurons)
# Unclear excluded from domain counting

progenitor_domains <- c("RP", "FP",
                        "dp1", "dp2", "dp3", "dp4/5", "dp6",
                        "p0",  "p1",  "p2",  "pMN",   "p3")
neurons_hh7_hh18     <- c("MN", "V2", "V3", "V0/V1", "dI1", "dI4/dI5")

# ── Build per-clone domain summary (multicell clones only) ────────────────────
clone_size_lookup <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

clone_domains <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  left_join(clone_size_lookup, by = "exp_clone") %>%
  filter(clone_size > 1) %>%
  mutate(broad_type = case_when(
    cell_type_collapsed %in% progenitor_domains ~ "progenitor",
    cell_type_collapsed %in% neurons_hh7_hh18     ~ "neuron",
    TRUE                                        ~ "excluded"   # Unclear
  )) %>%
  group_by(exp_clone) %>%
  summarise(
    prog_domain_list = list(sort(unique(cell_type_collapsed[broad_type == "progenitor"]))),
    n_prog_domains   = n_distinct(cell_type_collapsed[broad_type == "progenitor"]),
    contains_neurons = any(broad_type == "neuron"),
    .groups = "drop"
  ) %>%
  mutate(
    domain_fate = case_when(
      n_prog_domains == 0 & contains_neurons  ~ "neuron_only",
      n_prog_domains == 0 & !contains_neurons ~ "excluded_only",
      n_prog_domains == 1                     ~ "single_domain",
      n_prog_domains >= 2                     ~ "multi_domain"
    )
  )

cat("Total multicell clones:", nrow(clone_domains), "\n")

# ── Domain fate overall ───────────────────────────────────────────────────────
cat("\n=== Progenitor domain diversity (all multicell clones) ===\n")
domain_fate_overall <- clone_domains %>%
  filter(domain_fate != "excluded_only") %>%
  count(domain_fate) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(factor(domain_fate, levels = c("single_domain", "multi_domain", "neuron_only")))
print(domain_fate_overall)

# ── Multi-domain combos ───────────────────────────────────────────────────────
cat("\n=== Multi-domain combos ===\n")
multi_domain_combos <- clone_domains %>%
  filter(domain_fate == "multi_domain") %>%
  mutate(combo = sapply(prog_domain_list, function(x) paste(x, collapse = " + "))) %>%
  count(combo) %>%
  arrange(desc(n))
print(multi_domain_combos, n = 30)

# ── Save as multi-sheet Excel ─────────────────────────────────────────────────
wb2 <- createWorkbook()
addWorksheet(wb2, "domain_fate_overall")
writeData(wb2, "domain_fate_overall", domain_fate_overall)
addWorksheet(wb2, "multi_domain_combos")
writeData(wb2, "multi_domain_combos", multi_domain_combos)
saveWorkbook(wb2, file.path(data, "Table2_clone_domain_diversity_injHH7-collHH18.xlsx"), overwrite = TRUE)
cat("Table 2 saved!\n")
```

### Dropout fractions per sample
```{r dropout-fractions}
# Compute LARRY detection rates per cell type (per sample and overall); save as CSVs
celltype_order_hh7_hh18 <- c(
  "RP", "dp1", "dp2", "dp3", "dp4/5", "dp6",
  "p0", "p1", "p2", "pMN", "p3", "FP",
  "V0/V1", "dI1", "dI4/dI5", "V2", "V3", "MN",
  "Unclear"
)

# ── Per sample ────────────────────────────────────────────────────────────────
dropout_per_sample <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(cell_type_collapsed)) %>%
  mutate(cell_type_collapsed = factor(cell_type_collapsed, levels = celltype_order_hh7_hh18)) %>%
  filter(!is.na(cell_type_collapsed)) %>%           # drop values not in celltype_order_hh7_hh18
  group_by(sample, cell_type_collapsed, .drop = FALSE) %>%
  summarise(
    detected     = sum(clone_id_detected == "detected",     na.rm = TRUE),
    not_detected = sum(clone_id_detected == "not_detected", na.rm = TRUE),
    total_cells  = n(),
    .groups = "drop"
  ) %>%
  mutate(pct_detected = round(100 * detected / total_cells, 1)) %>%
  arrange(sample, cell_type_collapsed)

# ── Overall (across all LARRY samples) ────────────────────────────────────────
dropout_overall <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(cell_type_collapsed)) %>%
  mutate(cell_type_collapsed = factor(cell_type_collapsed, levels = celltype_order_hh7_hh18)) %>%
  filter(!is.na(cell_type_collapsed)) %>%           # drop values not in celltype_order_hh7_hh18
  group_by(cell_type_collapsed, .drop = FALSE) %>%
  summarise(
    detected     = sum(clone_id_detected == "detected",     na.rm = TRUE),
    not_detected = sum(clone_id_detected == "not_detected", na.rm = TRUE),
    total_cells  = n(),
    .groups = "drop"
  ) %>%
  mutate(pct_detected = round(100 * detected / total_cells, 1)) %>%
  arrange(cell_type_collapsed)

print(dropout_per_sample, n = Inf)
print(dropout_overall)

write.csv(dropout_per_sample, file.path(data, "LARRY_detection_summary_per_sample_injHH7-collHH18.csv"), row.names = FALSE)
write.csv(dropout_overall,    file.path(data, "LARRY_detection_summary_overall_injHH7-collHH18.csv"),    row.names = FALSE)
cat("Dropout fractions saved!\n")
```

### LARRY detection barplot
```{r detection-barplot-hh7-hh18}
# Stacked detected / not_detected, one bar for the whole dataset
detection_plot_df <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  mutate(clone_id_detected = ifelse(!is.na(exp_clone), "detected", "not_detected")) %>%
  group_by(clone_id_detected) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    percent_cells     = round(100 * n / sum(n), 1),
    dataset           = "HH7-9 → HH18",
    clone_id_detected = factor(clone_id_detected, levels = c("not_detected", "detected"))
  ) %>%
  arrange(clone_id_detected) %>%                      # not_detected first → bottom
  mutate(y_label = cumsum(percent_cells) - percent_cells / 2)

ggplot(detection_plot_df, aes(x = dataset, y = percent_cells, fill = clone_id_detected)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    aes(label = round(percent_cells, digits = 1), y = y_label),
    colour = "black", size = 8, hjust = 0.5, vjust = 0.5
  ) +
  scale_fill_manual(
    values = c("not_detected" = "grey90", "detected" = "darkcyan"),
    labels = c("not_detected" = "Not detected", "detected" = "Detected")
  ) +
  labs(title = "LARRY barcode detection") +
  ylab("% of cells") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    plot.title   = element_text(face = "bold", size = 24, vjust = 0.5, hjust = 0.5),
    axis.title.y = element_text(face = "bold", size = 18),
    axis.text    = element_text(size = 14)
  )

ggsave(filename = "LARRY_detection_barplot.pdf", plot = last_plot(), path = plots,
       width = 3, height = 12, units = "cm", dpi = 300, bg = "transparent")
```

---

## Export for Python h5ad reconstruction (run once)
```{r export-h5ad, eval=FALSE}
# Produces flat files → copy export_dir to NEMO before running scLiTr notebook:
#   scp -r <data>/h5ad_export/ boeziog@login.nemo.crick.ac.uk:<nemo_data>/injHH7-collHH18_LARRY/data/h5ad_export/

export_dir <- file.path(data, "h5ad_export")
dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

# 1. Sparse count matrix (genes × cells — Python transposes to cells × genes)
Matrix::writeMM(
  GetAssayData(seur_obj, assay = "RNA", layer = "counts"),
  file = file.path(export_dir, "counts.mtx")
)

# 2. Gene names and barcodes (same order as count matrix)
writeLines(rownames(seur_obj), file.path(export_dir, "genes.txt"))
writeLines(colnames(seur_obj), file.path(export_dir, "barcodes.txt"))

# 3. Metadata
seur_obj@meta.data %>%
  tibble::rownames_to_column("index") %>%
  select(index, clone = exp_clone, somites,
         cell_type_fine, cell_type_coarse, cell_type_collapsed) %>%
  write.csv(file.path(export_dir, "metadata.csv"), row.names = FALSE)

# 4. scVI embedding (cells × 30)
as.data.frame(seur_obj@reductions$scvi@cell.embeddings) %>%
  tibble::rownames_to_column("index") %>%
  write.csv(file.path(export_dir, "X_scvi.csv"), row.names = FALSE)

# 5. UMAP embedding (cells × 3)
as.data.frame(seur_obj@reductions$umap@cell.embeddings) %>%
  tibble::rownames_to_column("index") %>%
  write.csv(file.path(export_dir, "X_umap.csv"), row.names = FALSE)

cat("Export complete. Files in:", export_dir, "\n")
cat("Copy to NEMO before running scLiTr notebook.\n")
```
