---
title: "convert_anndata_seurat"
output: html_document
date: '2025-04-24'
---

This notebook converts the exported AnnData matrices from Python preprocessing into a normalised Seurat object for the chicken spinal cord LARRY dataset, so that downstream R analyses can use a single RDS file rather than raw flat files.

**Data required:**

- `preprocessing_larry_chicken_matrix.mtx` — sparse count matrix (cells × genes)
- `preprocessing_larry_chicken_genes.csv` — gene names
- `preprocessing_larry_chicken_cells.csv` — cell barcodes
- `preprocessing_larry_chicken_meta.csv` — per-cell metadata (includes `wave`, `cell_type`, etc.)
- `preprocessing_larry_chicken_umap.csv` — pre-computed 3-D UMAP coordinates

**Outputs:**

- `preprocessing_larry_chicken_clusters_newwave.rds` — normalised, scaled Seurat object with UMAP embedding and metadata attached

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r libraries}
# Load required packages for Seurat object creation, data wrangling, and sparse matrix I/O
library(Seurat)
library(tidyverse)
library(Matrix)
```

## Paths

```{r paths}
# Define input (anndata_seurat export) and output (seurat RDS) directories
anndata_seurat_path = "/camp/home/depottj/working/depottj/projects/briscoej/giulia.boezio/LARRY_spinal_cord/data/anndata_seurat"
seurat_path = "/camp/home/depottj/working/depottj/projects/briscoej/giulia.boezio/LARRY_spinal_cord/data/seurat"
```

## Make Seurat object

```{r make_seurat_object}
# Read count matrix, gene names, and cell barcodes then assemble a Seurat object
file_name = "chicken"

#Read in Count matrix
counts <- readMM(file.path(anndata_seurat_path , str_c("preprocessing_larry_" , file_name , "_matrix.mtx")))

#Read in gene names
gene_names <- read.csv(file.path(anndata_seurat_path , str_c("preprocessing_larry_" , file_name , "_genes.csv")))[["Gene"]]

#Read in cell names
cell_names <- read.csv(file.path(anndata_seurat_path , str_c("preprocessing_larry_" , file_name , "_cells.csv")))[["Cells"]]


colnames(counts) <- gene_names
rownames(counts) <- cell_names

seurat_obj <- CreateSeuratObject(counts = t(counts))
```

```{r add_metadata}
# Read per-cell metadata from CSV and attach it to the Seurat object
meta_data <- read_csv(file.path(anndata_seurat_path , str_c("preprocessing_larry_" , file_name , "_meta.csv"))) %>%
              column_to_rownames(var = "index")

seurat_obj@meta.data <- meta_data
```

```{r normalise_scale}
# Log-normalise counts and z-score scale gene expression across cells
seurat_obj_norm_scale <- seurat_obj %>%
  NormalizeData(normalization.method = "LogNormalize" , scale.factor = 10000) %>%
  ScaleData()
```

```{r umap}
# Import pre-computed 3-D UMAP coordinates from Python and embed them in the Seurat object
umap_coord <- read_csv(file.path(anndata_seurat_path , str_c("preprocessing_larry_" , file_name , "_umap.csv")), col_names = FALSE)
colnames(umap_coord) <- c("UMAP_1","UMAP_2","UMAP_3")
rownames(umap_coord) <- cell_names

#Make umap to matrix
umap_matrix <- as.matrix(umap_coord)

#Create Reduction object
umap_reduc <- CreateDimReducObject(embeddings = umap_matrix, key = "UMAP_", assay = DefaultAssay(seurat_obj_norm_scale))

#Add umap to Seurat object
seurat_obj_norm_scale[["umap"]] <- umap_reduc

DimPlot(seurat_obj_norm_scale, reduction = "umap", group.by = "wave")
DimPlot(seurat_obj_norm_scale, reduction = "umap", group.by = "cell_type")
```

```{r write_output}
# Save the finalised Seurat object to disk as an RDS file
saveRDS(seurat_obj_norm_scale, file = file.path(seurat_path , str_c("preprocessing_larry_" , file_name , "_clusters_newwave.rds")))

```
