---
title: "Figure_4_plots"
output: html_document
date: "2025-01-29"
---

This notebook generates Figure 6 and Supplementary Figures S7, S8, and S9 for the injHH12-collHH28/31/35 chicken spinal cord LARRY dataset (human CS16/CS17 samples), covering dI4/dI5 clonal classification, differential gene expression, clone QC tables, LARRY detection barplots, UMAPs, Spearman correlation with the Rayon in vivo reference, and binomial mixture modelling with bootstrapping.

**Data required:**
- `preprocessing_larry_human_clusters.rds` — integrated human LARRY Seurat object
- `Rayon_Vs_Slice_Timepoint_Spearman_Correlations.csv` — precomputed Spearman correlations between pseudobulk slice samples and the Rayon 2021 in vivo reference

**Outputs:**
- `volcano-DEGs-pval.pdf` — volcano plot of DEGs between dI4 cells sharing/not sharing with dI5
- `degs_dI4_human.csv` — DEG table for dI4 (shared vs dI4-only)
- `degs_dI5_human.csv` — DEG table for dI5 (shared vs dI5-only)
- `Table1_clone_QC_human.xlsx` — per-sample clone QC summary table
- `clone_size_histogram.pdf` — clone size distribution histogram
- `Table2_clone_composition_neuronal_fate_human.xlsx` — clone composition and neuronal fate tables
- `LARRY_detection_summary_per_sample_human.csv` / `_overall_human.csv` — dropout fraction tables
- `LARRY_detection_barplot.pdf` — stacked barplot of LARRY barcode detection rate (Fig. S8)
- `umap-sub-colors-fine-degrading-legend-dim1-3.png` — cell-type UMAP (Fig. 6)
- `Spearman_corr_slices_Rayon.pdf` — correlation heatmap vs Rayon reference
- `dI4_dI5_UMAP-dim1-3.png` — UMAP highlighting dI4/dI5 populations
- `dI4_dI5_ratio_all_median_replicates.pdf` — stacked median ratio bar chart by experiment
- `dI4_dI5_KDE_tot.pdf` — KDE/histogram of dI4 fraction in clones
- `UMAP_samples_colors-nolegend.png` / `_split.png` — UMAPs coloured by experiment (Fig. S7)
- `clone_detected-noledeng.png` / `multicellular_clones-nolegend.png` — detection and multi-cell clone UMAPs (Fig. S8)
- `multicell_clones_cells_pie.pdf` / `multicell_vs_single_clones_count_pie.pdf` / `clone_detected_avg_pie.pdf` / `clone_detected_experiments_pie.pdf` — pie charts of clone statistics (Fig. S8)
- `Upset_plots_broad_cat.pdf` — UpSet plot of neuron/progenitor clonal sharing
- `top10_clones_dI4-dI5shared.png` — UMAP of top 10 largest shared dI4/dI5 clones
- `n_cells_all_clones_vln.pdf` / `n_cells_multicell_clones_vln.pdf` — violin plots of clone size
- `barplot_observed_vs_binomial.pdf` — observed vs binomial expected clone counts (Fig. 6I)
- `fixed_model_alpha_loglikelihood.pdf` — profile log-likelihood for alpha parameter (Fig. S9A)
- `chi_square_distributions_models_nofree.pdf` — bootstrap chi-square distributions (Fig. S9B)
- `fit_summary_metrics_models_nofree.csv` / `models_metrics_nofree.csv` / `Table2-bootstrapped_mean_bin_counts_comparison_nofree.csv` — model comparison tables

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r init}
library(dplyr)
library(Seurat)
library(readr)
library(tidyverse)
library(scCustomize)
library(ggplot2)
library(ggrepel)
library(glue)

```

## Paths
```{r define-paths}
# Set output directories for plots and data

plots = "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/plots/new_integrated/paper/figure6/"

data = "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/data/"

```

## Read RDS
```{r load-seurat-object}
# Loads the human LARRY Seurat object, standardises metadata column names, assigns experiment labels, and computes multi-cell clone status

seur_obj <- read_rds("/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/human_LARRY/SO/preprocessing_larry_human_clusters.rds")


colnames(seur_obj@meta.data)[colnames(seur_obj@meta.data) == "cell_type"] <- "Cell.type"

colnames(seur_obj@meta.data)[colnames(seur_obj@meta.data) == "clone"] <- "exp_clone"

# Column 1: "experiment"
seur_obj@meta.data$experiment <- case_when(
  seur_obj@meta.data$sample %in% c("BOE7075A1", "BOE7075A2") ~ "CS17",
  seur_obj@meta.data$sample %in% c("BOE7075A4") ~ "CS16",
  TRUE ~ NA_character_
)


# Column 2: "samples_LARRY"
seur_obj@meta.data$samples_LARRY <- case_when(
  seur_obj@meta.data$sample %in% c("BOE7075A2", "BOE7075A4") ~ "LARRY",
  seur_obj@meta.data$sample %in% c("BOE7075A1") ~ "control",
  TRUE ~ NA_character_
)

seur_obj@meta.data <- seur_obj@meta.data %>%
  mutate(clone_id_detected = case_when(
    samples_LARRY %in% c("LARRY") & !is.na(exp_clone) ~ "detected",
   samples_LARRY %in% c("LARRY") & is.na(exp_clone) ~ "not_detected",
    TRUE ~ NA_character_
  ))

seur_obj@meta.data$Cell.type <- seur_obj@meta.data$Cell.type %>%
  recode("Immature neurons" = "immature neurons") %>%
  replace_na("Unknown")


### Clone size
# Step 1: Compute clone sizes from exp_clone
clone_sizes <- seur_obj@meta.data %>%
  filter(!is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

# Step 2: Join clone sizes back to meta.data safely
meta_with_clone_status <- seur_obj@meta.data %>%
  left_join(clone_sizes, by = "exp_clone") %>%
  mutate(multi_cell_clone = case_when(
    clone_size > 1 ~ "multi-cell_clone",
    clone_size == 1 ~ "single-cell_clone",
    TRUE ~ NA_character_  # keep NA where no exp_clone
  ))

# Ensure rownames are preserved
rownames(meta_with_clone_status) <- rownames(seur_obj@meta.data)

# Step 4: Add the new column to the Seurat object
seur_obj <- AddMetaData(seur_obj, metadata = meta_with_clone_status$multi_cell_clone, col.name = "multi_cell_clone")


#Count number unique clones
filtered_n_clones <- seur_obj@meta.data %>%
  filter(!is.na(exp_clone),
         multi_cell_clone == "multi-cell_clone") %>%
  pull(exp_clone) %>%
  unique() %>%
  length()

filtered_n_clones

```

## Divide dI4 into clones that share with dI5 and those that do not
```{r classify-dI4-sharing}
# Classifies dI4_dILA cells in multi-cell clones as either sharing a clone with dI5_dILB or dI4-only

# Step 1: Filter for dI4_dILA and dI5_dILB cells, keeping the "experiment" column
dI4_cells <- seur_obj@meta.data %>%
  as_tibble(rownames = "cell_id") %>%  # Convert row names (cell IDs) into a "cell_id" column
  filter(Cell.type == "dI4_dILA", !is.na(exp_clone), multi_cell_clone == "multi-cell_clone") %>%  # dI4_dILA cells
  dplyr::select(cell_id, exp_clone, experiment, Cell.type)  # Keep relevant columns including experiment

dI5_cells <- seur_obj@meta.data %>%
  as_tibble(rownames = "cell_id") %>%  # Convert row names into a "cell_id" column
  filter(Cell.type == "dI5_dILB", !is.na(exp_clone), multi_cell_clone == "multi-cell_clone") %>%  # dI5_dILB cells
  dplyr::select(exp_clone, cell_id)  # We only need the exp_clone column from dI5 cells

# Step 2: Identify shared clonal groups (exp_clone) between dI4_dILA and dI5_dILB
shared_clones <- intersect(dI4_cells$exp_clone, dI5_cells$exp_clone)

# Step 3: Categorize dI4_dILA cells based on shared exp_clone, keeping the experiment column
dI4_cells_classes <- dI4_cells %>%
  mutate(
    clone_category = ifelse(exp_clone %in% shared_clones,
                            "shared_with_dI5",
                            "dI4_only")
  )
```



### Check for differential gene expression between groups
```{r add-dI4-category-to-seurat}
# Adds the dI4_clone_category label to Seurat metadata and subsets the object to dI4 cells

# Add clone_category information to the Seurat object
seur_obj@meta.data$dI4_clone_category <- NA  # Create an empty column

# Assign the clone categories from dI4_cells_classes to the Seurat object
seur_obj@meta.data[dI4_cells_classes$cell_id, "dI4_clone_category"] <- dI4_cells_classes$clone_category

# df = seur_obj@meta.data$dI4_clone_category

# Subset the Seurat object to keep only dI4_dILA cells
seur_obj_dI4 <- subset(seur_obj, cells = dI4_cells$cell_id)

# seur_obj_dI4_HH28 <- subset(seur_obj,
#                              cells = dI4_cells$cell_id[dI4_cells$experiment == "HH28"])



#Plot diff cells on umap
# Convert to character first
seur_obj@meta.data$dI4_clone_category <- as.character(seur_obj@meta.data$dI4_clone_category)

# Replace NA with "Unknown"
seur_obj@meta.data$dI4_clone_category[is.na(seur_obj@meta.data$dI4_clone_category)] <- "Unknown"

# Re-factor with desired level order
seur_obj@meta.data$dI4_clone_category <- factor(
  seur_obj@meta.data$dI4_clone_category,
  levels = c("Unknown", "dI4_only","shared_with_dI5"))


```


## Calculate DEGs with Seurat
```{r DEG-dI4-volcano}
# Runs FindMarkers between dI4 shared-with-dI5 and dI4-only cells, then produces a volcano plot

# 1. Differential expression analysis
diff_exp_results <- FindMarkers(
  seur_obj_dI4,
  ident.1 = "shared_with_dI5",
  ident.2 = "dI4_only",
  group.by = "dI4_clone_category",
  min.pct = 0.1
)

# 2. Annotate significance and labels
diff_exp_results <- diff_exp_results %>%
  rownames_to_column(var = "gene_id") %>%
  mutate(
    significant = case_when(
      p_val < 0.05 & avg_log2FC > 0.5  ~ "Upregulated",
      p_val < 0.05 & avg_log2FC < -0.5 ~ "Downregulated",
      TRUE ~ "Not significant"
    )
  )

# Top 30 significant genes by p-value
top30 <- diff_exp_results %>%
  filter(significant != "Not significant") %>%
  slice_min(order_by = p_val, n = 30)

# 3. Volcano plot
p <- ggplot(diff_exp_results, aes(x = avg_log2FC, y = -log10(p_val))) +
  geom_point(aes(color = significant), size = 2, alpha = 0.6) +
  geom_text_repel(
    data = top30,
    aes(label = gene_id),
    size = 3,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = c(
    "Upregulated"     = "darkred",
    "Downregulated"   = "darkblue",
    "Not significant" = "grey85"
  )) +
  labs(
    title = "Genes up and down in dI4 sharing with dI5 vs dI4-only cells",
    x = "Log2 Fold Change",
    y = "-log10(p-value)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

# 4. Save plot
ggsave(filename = "volcano-DEGs-pval.pdf", plot = p, path = plots,
       scale = 1, width = 20, height = 14, units = "cm", dpi = 300, bg = "transparent")

```

```{r DEG-dI4-save-csv}
# Reruns FindMarkers without logfc threshold and saves the full DEG table for dI4 cells to CSV

# Subset to dI4 cells only
seur_dI4 <- subset(seur_obj, cells = rownames(seur_obj@meta.data)[
  seur_obj@meta.data$dI4_clone_category %in% c("shared_with_dI5", "dI4_only")
])

# Run DEGs: positive log2FC = higher in shared_with_dI5
degs <- FindMarkers(
  seur_dI4,
  ident.1 = "shared_with_dI5",
  ident.2 = "dI4_only",
  group.by = "dI4_clone_category",
  min.pct = 0.1,
  logfc.threshold = 0
)

# Clean output
degs_out <- degs %>%
  rownames_to_column("gene") %>%
  dplyr::select(gene, avg_log2FC, p_val, p_val_adj, pct.1, pct.2) %>%
  arrange(p_val)

# Save — run once per species with appropriate filename
write.csv(degs_out, "degs_dI4_human.csv", row.names = FALSE)
# or write.csv(degs_out, "degs_dI4_human.csv", row.names = FALSE)
```

## dI5 DEGs
```{r DEG-dI5-save-csv}
# Classifies dI5_dILB cells as sharing with dI4 or dI5-only, runs FindMarkers, and saves the DEG table to CSV

# Step 1: dI5_dILB cells with all needed columns (note: original dI5_cells only kept exp_clone + cell_id)
dI5_cells <- seur_obj@meta.data %>%
  as_tibble(rownames = "cell_id") %>%
  filter(Cell.type == "dI5_dILB", !is.na(exp_clone), multi_cell_clone == "multi-cell_clone") %>%
  dplyr::select(cell_id, exp_clone, experiment, Cell.type)

dI4_cells_ref <- seur_obj@meta.data %>%
  as_tibble(rownames = "cell_id") %>%
  filter(Cell.type == "dI4_dILA", !is.na(exp_clone), multi_cell_clone == "multi-cell_clone") %>%
  dplyr::select(exp_clone, cell_id)

# Step 2: Shared clones between dI5_dILB and dI4_dILA
shared_clones_dI5 <- intersect(dI5_cells$exp_clone, dI4_cells_ref$exp_clone)

# Step 3: Categorise dI5 cells
dI5_cells_classes <- dI5_cells %>%
  mutate(
    clone_category = ifelse(exp_clone %in% shared_clones_dI5,
                            "shared_with_dI4",
                            "dI5_only")
  )

# Step 4: Add to Seurat metadata
seur_obj@meta.data$dI5_clone_category <- NA
seur_obj@meta.data[dI5_cells_classes$cell_id, "dI5_clone_category"] <- dI5_cells_classes$clone_category

seur_obj@meta.data$dI5_clone_category <- as.character(seur_obj@meta.data$dI5_clone_category)
seur_obj@meta.data$dI5_clone_category[is.na(seur_obj@meta.data$dI5_clone_category)] <- "Unknown"
seur_obj@meta.data$dI5_clone_category <- factor(
  seur_obj@meta.data$dI5_clone_category,
  levels = c("Unknown", "dI5_only", "shared_with_dI4")
)

# Step 5: Subset and run DEGs
seur_dI5 <- subset(seur_obj, cells = rownames(seur_obj@meta.data)[
  seur_obj@meta.data$dI5_clone_category %in% c("shared_with_dI4", "dI5_only")
])

# positive log2FC = higher in shared_with_dI4
degs_dI5 <- FindMarkers(
  seur_dI5,
  ident.1 = "shared_with_dI4",
  ident.2 = "dI5_only",
  group.by = "dI5_clone_category",
  min.pct = 0.1,
  logfc.threshold = 0
)

degs_dI5_out <- degs_dI5 %>%
  rownames_to_column("gene") %>%
  dplyr::select(gene, avg_log2FC, p_val, p_val_adj, pct.1, pct.2) %>%
  arrange(p_val)


write.csv(degs_dI5_out, "degs_dI5_human.csv", row.names = FALSE)
```



## Clone QC, composition and neuronal fate

### Clone QC — Table 1
```{r clone-qc-table}
# Computes per-sample and overall clone QC statistics (clone counts, size distribution, detection rates) and saves to Excel
library(tidyverse)
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
library(openxlsx)

data_path <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/data/figure6"
if (!dir.exists(data_path)) dir.create(data_path, recursive = TRUE)

# ── Per-sample clone stats ────────────────────────────────────────────────────
clone_sizes_raw <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, sample, name = "clone_size")

detection_per_sample <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  group_by(sample) %>%
  summarise(
    total_cells          = n(),
    cells_LARRY_detected = sum(clone_id_detected == "detected", na.rm = TRUE),
    pct_detected         = round(100 * cells_LARRY_detected / total_cells, 1),
    .groups = "drop"
  )

clones_per_sample <- clone_sizes_raw %>%
  group_by(sample) %>%
  summarise(
    n_clones      = n_distinct(exp_clone),
    mean_size     = round(mean(clone_size), 2),
    median_size   = median(clone_size),
    max_size      = max(clone_size),
    n_single_cell = sum(clone_size == 1),
    n_multi_cell  = sum(clone_size > 1),
    pct_single    = round(100 * sum(clone_size == 1) / n(), 1),
    pct_multi     = round(100 * sum(clone_size > 1)  / n(), 1),
    .groups = "drop"
  ) %>%
  left_join(detection_per_sample, by = "sample")

# ── Overall row ───────────────────────────────────────────────────────────────
clone_sizes_all <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

larry_cells_all <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY")

table1_overall <- tibble(
  sample               = "Overall",
  n_clones             = nrow(clone_sizes_all),
  mean_size            = round(mean(clone_sizes_all$clone_size), 2),
  median_size          = median(clone_sizes_all$clone_size),
  max_size             = max(clone_sizes_all$clone_size),
  n_single_cell        = sum(clone_sizes_all$clone_size == 1),
  n_multi_cell         = sum(clone_sizes_all$clone_size > 1),
  pct_single           = round(100 * sum(clone_sizes_all$clone_size == 1) / nrow(clone_sizes_all), 1),
  pct_multi            = round(100 * sum(clone_sizes_all$clone_size > 1)  / nrow(clone_sizes_all), 1),
  total_cells          = nrow(larry_cells_all),
  cells_LARRY_detected = sum(larry_cells_all$clone_id_detected == "detected", na.rm = TRUE),
  pct_detected         = round(100 * sum(larry_cells_all$clone_id_detected == "detected", na.rm = TRUE) /
                                 nrow(larry_cells_all), 1)
)

table1 <- bind_rows(clones_per_sample, table1_overall)
print(table1)

wb1 <- createWorkbook()
addWorksheet(wb1, "summary_per_sample")
writeData(wb1, "summary_per_sample", table1)
saveWorkbook(wb1, file.path(data_path, "Table1_clone_QC_human.xlsx"), overwrite = TRUE)
cat("Table 1 saved!\n")
```

### Clone size histogram
```{r clone-size-histogram}
# Plots a histogram of clone size distribution across all LARRY clones with median and mean annotations

plots_fig6 <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/plots/new_integrated/paper/figure6/"

clone_sizes_hist <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

annot <- clone_sizes_hist %>%
  summarise(med = median(clone_size), mean = round(mean(clone_size), 2), n = n())

ggplot(clone_sizes_hist, aes(x = clone_size)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = annot$med, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("n = ", annot$n, "\nmedian = ", annot$med, "\nmean = ", annot$mean),
           hjust = 1.1, vjust = 1.2, size = 3.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  theme_minimal(base_size = 13) +
  labs(title = "Clone size distribution (all LARRY clones — human)",
       x = "Clone size (number of cells)", y = "Number of clones")

ggsave(filename = "clone_size_histogram.pdf", plot = last_plot(), path = plots_fig6,
       scale = 1, width = 20, height = 15, units = "cm", dpi = 300, bg = "transparent")
```

### Clone composition and neuronal fate — Table 2
```{r clone-composition-fate}
# Classifies multi-cell clones by broad fate (progenitor/neuron/glia) and neuronal fate diversity, saves results to Excel

# ── Cell type categories (human) ──────────────────────────────────────────────
progenitors <- c("RP-dp1", "dp1-3", "dpL", "dp6/p2", "pMN/p3", "FP")
neurons      <- c("dI1", "dI2/3", "dI4_dILA", "dI5_dILB", "dI6", "V0", "V1", "V2", "MN", "V3")
glia         <- c("OPCs")
# Unknown and mesoderm/drg → excluded from broad_type counting

clone_size_lookup <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  count(exp_clone, name = "clone_size")

# ── Build clone-level fate (multicell only) ───────────────────────────────────
clone_fate <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(exp_clone)) %>%
  left_join(clone_size_lookup, by = "exp_clone") %>%
  filter(clone_size > 1) %>%
  mutate(broad_type = case_when(
    Cell.type %in% progenitors ~ "progenitor",
    Cell.type %in% neurons     ~ "neuron",
    Cell.type %in% glia        ~ "glia",
    TRUE                       ~ "excluded"
  )) %>%
  group_by(exp_clone) %>%
  summarise(
    broad_types    = list(unique(broad_type[broad_type != "excluded"])),
    neuron_types   = list(unique(Cell.type[broad_type == "neuron"])),
    n_neuron_types = n_distinct(Cell.type[broad_type == "neuron"]),
    .groups = "drop"
  ) %>%
  mutate(
    clone_class = case_when(
      sapply(broad_types, length) == 0                                                              ~ "excluded_only",
      sapply(broad_types, function(x) setequal(x, "progenitor"))                                   ~ "progenitor_only",
      sapply(broad_types, function(x) setequal(x, "neuron"))                                       ~ "neuron_only",
      sapply(broad_types, function(x) setequal(x, "glia"))                                         ~ "glia_only",
      sapply(broad_types, function(x) all(c("progenitor","neuron") %in% x) & !"glia" %in% x)      ~ "progenitor_neuron",
      sapply(broad_types, function(x) all(c("glia","neuron") %in% x) & !"progenitor" %in% x)      ~ "glia_neuron",
      sapply(broad_types, function(x) all(c("glia","progenitor") %in% x) & !"neuron" %in% x)      ~ "glia_progenitor",
      TRUE                                                                                           ~ "mixed_all"
    ),
    neuron_fate = case_when(
      n_neuron_types == 0 ~ NA_character_,
      n_neuron_types == 1 ~ "single_neuronal_fate",
      n_neuron_types >  1 ~ "multi_neuronal_fate"
    )
  )

cat("Total multicell clones:", nrow(clone_fate), "\n")

# ── Clone class composition ───────────────────────────────────────────────────
cat("\n=== Clone composition overall ===\n")
clone_class_overall <- clone_fate %>%
  filter(clone_class != "excluded_only") %>%
  count(clone_class) %>%
  mutate(pct = round(100 * n / sum(n), 1))
print(clone_class_overall)

# ── Neuronal fate (clones with neurons as denominator) ────────────────────────
cat("\n=== Neuronal fate (clones with neurons) ===\n")
fate_overall <- clone_fate %>%
  filter(!is.na(neuron_fate)) %>%
  count(neuron_fate) %>%
  mutate(pct = round(100 * n / sum(n), 1))
print(fate_overall)

# ── Neuronal fate (all multicell clones as denominator) ───────────────────────
cat("\n=== Neuronal fate — ALL multicell clones ===\n")
fate_all_overall <- clone_fate %>%
  mutate(fate_category = case_when(
    is.na(neuron_fate) ~ "no_neuronal_fate",
    TRUE               ~ neuron_fate
  )) %>%
  count(fate_category) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(factor(fate_category, levels = c("no_neuronal_fate",
                                           "single_neuronal_fate",
                                           "multi_neuronal_fate")))
print(fate_all_overall)

# ── Multi-fate combos ─────────────────────────────────────────────────────────
cat("\n=== Multi-fate combos ===\n")
multi_fate_combos <- clone_fate %>%
  filter(n_neuron_types > 1) %>%
  mutate(combo = sapply(neuron_types, function(x) paste(sort(x), collapse = " + "))) %>%
  count(combo) %>%
  arrange(desc(n))
print(multi_fate_combos, n = 30)

# ── Fate by neuron type ───────────────────────────────────────────────────────
fate_by_type <- clone_fate %>%
  filter(!is.na(neuron_fate)) %>%
  tidyr::unnest(neuron_types) %>%
  rename(neuron_type = neuron_types) %>%
  group_by(neuron_type) %>%
  summarise(
    total_clones  = n(),
    n_multi_fate  = sum(neuron_fate == "multi_neuronal_fate"),
    n_single_fate = sum(neuron_fate == "single_neuronal_fate"),
    pct_multi     = round(100 * n_multi_fate  / total_clones, 1),
    pct_single    = round(100 * n_single_fate / total_clones, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_multi))

cat("\n=== Neuronal fate by type ===\n")
print(fate_by_type)

# ── Save as multi-sheet Excel ─────────────────────────────────────────────────
wb2 <- createWorkbook()
addWorksheet(wb2, "clone_class_overall")
writeData(wb2, "clone_class_overall", clone_class_overall)
addWorksheet(wb2, "fate_overall")
writeData(wb2, "fate_overall", fate_overall)
addWorksheet(wb2, "fate_all_clones_overall")
writeData(wb2, "fate_all_clones_overall", fate_all_overall)
addWorksheet(wb2, "fate_by_type")
writeData(wb2, "fate_by_type", fate_by_type)
addWorksheet(wb2, "multi_fate_combos")
writeData(wb2, "multi_fate_combos", multi_fate_combos)
saveWorkbook(wb2, file.path(data_path, "Table2_clone_composition_neuronal_fate_human.xlsx"), overwrite = TRUE)
cat("Table 2 saved!\n")
```

## Dropout fractions
```{r dropout-fractions}
# Computes per-sample and overall LARRY detection fractions per cell type and saves to CSV

data_path <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/data/figure6"
if (!dir.exists(data_path)) dir.create(data_path, recursive = TRUE)

celltype_order <- c("RP-dp1","dp1-3", "dpL", "dp6/p2", "pMN/p3", "FP", "OPCs",
                    "dI1", "dI2/3","dI4_dILA","dI5_dILB", "dI6","V0","V1","V2","MN","V3",
                    "Unknown", "mesoderm/drg")

# ── Per sample ────────────────────────────────────────────────────────────────
dropout_per_sample <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(Cell.type)) %>%
  mutate(Cell.type = factor(Cell.type, levels = celltype_order)) %>%
  filter(!is.na(Cell.type)) %>%
  group_by(sample, Cell.type, .drop = FALSE) %>%
  summarise(
    detected     = sum(clone_id_detected == "detected",     na.rm = TRUE),
    not_detected = sum(clone_id_detected == "not_detected", na.rm = TRUE),
    total_cells  = n(),
    .groups = "drop"
  ) %>%
  mutate(pct_detected = round(100 * detected / total_cells, 1)) %>%
  arrange(sample, Cell.type)

# ── Overall ───────────────────────────────────────────────────────────────────
dropout_overall <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY", !is.na(Cell.type)) %>%
  mutate(Cell.type = factor(Cell.type, levels = celltype_order)) %>%
  filter(!is.na(Cell.type)) %>%
  group_by(Cell.type, .drop = FALSE) %>%
  summarise(
    detected     = sum(clone_id_detected == "detected",     na.rm = TRUE),
    not_detected = sum(clone_id_detected == "not_detected", na.rm = TRUE),
    total_cells  = n(),
    .groups = "drop"
  ) %>%
  mutate(pct_detected = round(100 * detected / total_cells, 1)) %>%
  arrange(Cell.type)

print(dropout_per_sample, n = Inf)
print(dropout_overall)

write.csv(dropout_per_sample, file.path(data_path, "LARRY_detection_summary_per_sample_human.csv"), row.names = FALSE)
write.csv(dropout_overall,    file.path(data_path, "LARRY_detection_summary_overall_human.csv"),    row.names = FALSE)
cat("Dropout fractions saved!\n")
```


### LARRY detection barplot
```{r detection-barplot-human}
# Plots a stacked barplot of LARRY barcode detection rate for the CS16-17 dataset

plots_fig6 <- "/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/SC22233/New-SO/plots/new_integrated/paper/figure6/"

# Stacked detected / not_detected, one bar for the whole dataset
detection_plot_df <- seur_obj@meta.data %>%
  filter(samples_LARRY == "LARRY") %>%
  mutate(clone_id_detected = ifelse(!is.na(exp_clone), "detected", "not_detected")) %>%
  group_by(clone_id_detected) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    percent_cells     = round(100 * n / sum(n), 1),
    dataset           = "CS16-17",
    clone_id_detected = factor(clone_id_detected, levels = c("not_detected", "detected"))
  ) %>%
  arrange(clone_id_detected) %>%                      # not_detected first → bottom
  mutate(y_label = cumsum(percent_cells) - percent_cells / 2)

ggplot(detection_plot_df, aes(x = dataset, y = percent_cells, fill = clone_id_detected)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    aes(label = round(percent_cells, digits = 1), y = y_label),
    colour = "black", size = 8, hjust = 0.5, vjust = 0.5
  ) +
  scale_fill_manual(
    values = c("not_detected" = "grey90", "detected" = "darkcyan"),
    labels = c("not_detected" = "Not detected", "detected" = "Detected")
  ) +
  labs(title = "LARRY barcode detection") +
  ylab("% of cells") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    plot.title   = element_text(face = "bold", size = 24, vjust = 0.5, hjust = 0.5),
    axis.title.y = element_text(face = "bold", size = 18),
    axis.text    = element_text(size = 14)
  )

ggsave(filename = "LARRY_detection_barplot.pdf", plot = last_plot(), path = plots_fig6,
       width = 3, height = 12, units = "cm", dpi = 300, bg = "transparent")
```



## Fig. 6
### Set new colours
```{r define-cell-type-colors}
# Defines the named colour vector for human cell types used in all Fig. 6 plots

celltype_order <- c("RP-dp1","dp1-3", "dpL", "dp6/p2", "pMN/p3",  "FP", "OPCs", "dI1" , "dI2/3","dI4_dILA","dI5_dILB", "dI6","V0","V1","V2","MN","V3", "Unknown", "mesoderm/drg")


colors_ids_fine <- c(
              "lightskyblue", #RP
                     "deepskyblue", # dp1_2
                       "deepskyblue3", #dpL
                     "deepskyblue4", # DP6-P2
              "dodgerblue4", #pMN-p3
              "blue", #FP
              "darkorchid4", # OPCs
              "wheat2",#dI1
                     "khaki", #dI2/3
       "goldenrod2", #dI4_dILA
              "darkgoldenrod1", #dI5_dILB
              "darkolivegreen3", #dI6
              "limegreen", #V0
              "palegreen3", # V1
              "darkseagreen", #V2
              "darkgreen", #MN
              "seagreen", #V3
       "grey95", #Unknwo
       "grey70" #meso
  )

# Create a named vector for colors
cell_type_colors <- setNames(colors_ids_fine, celltype_order)

```

### UMAPs
```{r umap-celltypes, fig.height=16, fig.width=20}
# Plots the full UMAP coloured by cell type using the defined colour palette (Fig. 6)

#Plot for normal UMAP with cell type names


# Ensure Cell.type follows the specified order
seur_obj$Cell.type <- factor(seur_obj$Cell.type, levels = celltype_order)
seur_obj <- SetIdent(seur_obj, value = seur_obj@meta.data$Cell.type)


#with degrading colours
DimPlot(seur_obj,
            label = F,
        pt.size = 1,
        repel = T,
        dims = c(1, 3)
        ) +
      # NoLegend() +
      coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL) +
      scale_color_manual(values = cell_type_colors) +
      guides(color = guide_legend(override.aes = list(size = 4))) + # Adjust legend appearance if needed
      theme(
        legend.position = "right",  # Forces legend to stay in a fixed position
        legend.key.size = unit(0.5, "cm"),  # Standardize legend key size
        legend.text = element_text(size = 12)  # Standardize legend text size
      )


ggsave(filename = "umap-sub-colors-fine-degrading-legend-dim1-3.png", plot = last_plot(), path = plots,
    scale = 1, width = 45 ,
    height = 40, units = "cm",
    dpi = 300, bg = "transparent")
```


### Spearman and Pearson correlation with in vivo
Use EntropySorting to find a set of 781 genes that separate cell states in the Rayon dataset (Rayon et al, 2021).
For each timepoint of the Rayon dataset create pseudobulk gene expression by averaging the expression of each of the 781 genes.
Do the same for each of the samples in the slice data.
Calculate the Pearsons correlations between pseudobulk samples.
```{r spearman-corr-heatmap}
# Loads precomputed Spearman correlations and plots them as a heatmap vs Rayon 2021 in vivo reference

# Load libraries
library(tidyverse)
library(readr)

# Load data
df <- read_csv("/Users/boeziog/LARRY_local/10X_analyses_local/Seurat/human_LARRY/Rayon_Vs_Slice_Timepoint_Spearman_Correlations.csv")

# Convert to long format
df_long <- df %>%
  pivot_longer(-1, names_to = "Slice_Timepoint", values_to = "Correlation") %>%
  rename(Rayon = 1)

# Plot heatmap
ggplot(df_long, aes(x = Slice_Timepoint, y = Rayon, fill = Correlation)) +
  geom_tile(color = "white") +
  # scale_fill_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B", midpoint = 0)+
scale_fill_viridis_c(option = "viridis")+
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(title = "Correlation Heatmap", x = NULL, y = NULL)

ggsave(
  filename = "Spearman_corr_slices_Rayon.pdf",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 20,
  height = 15,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)
```


### dI4/dI5 UMAP
```{r umap-dI4-dI5-highlight}
# Plots a UMAP highlighting dI4_dILA and dI5_dILB populations in distinct colours

# Define the clusters for each group
highlight_clusters_1 <- c("dI4_dILA")  # Replace with your actual cluster names
highlight_clusters_2 <- c("dI5_dILB")  # Replace with your actual cluster names

# Create a new metadata column for highlighting
seur_obj@meta.data$highlight <- ifelse(seur_obj@meta.data$`Cell.type` %in% highlight_clusters_1, "dI4_dILA",
                               ifelse(seur_obj@meta.data$`Cell.type` %in% highlight_clusters_2, "dI5_dILB", "Other"))

DimPlot(seur_obj, dim = c(1,3), group.by = "highlight", cols = c("dI4_dILA" = "#3F7D83", "dI5_dILB" = "#BA600F", "Other" = "grey"))+
  labs(title = NULL) +
      coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL)

#Save
ggsave(filename = paste0("dI4_dI5_UMAP-dim1-3", ".png"), plot = last_plot(), path = plots,scale = 1, width = 23 ,
    height = 20, units = "cm",
    dpi = 300, bg = "transparent")

```


### dI4/dI5 ratio — inverted median replicates
```{r dI4-dI5-ratio-stacked-bars}
# Computes per-replicate dI4/dI5 ratios and plots stacked median bars with SEM error bars by experiment

library(dplyr)
library(ggplot2)

# Define experiment order
experiment_order <- c("CS16", "CS17")

# Step 1: Compute per-replicate ratios
replicate_ratios <- seur_obj@meta.data %>%
  filter(
    Cell.type %in% c("dI4_dILA", "dI5_dILB")
    # ,
    # clone_id_detected == "detected"
  ) %>%
  mutate(experiment = factor(experiment, levels = experiment_order)) %>%
  group_by(experiment, sample, Cell.type) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(experiment, sample) %>%
  mutate(
    total = sum(n),
    ratio = n / total
  ) %>%
  ungroup()

# Step 2: Get median and SEM
summary_ratios <- replicate_ratios %>%
  group_by(experiment, Cell.type) %>%
  summarise(
    median_ratio = median(ratio),
    sem = sd(ratio) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    experiment = factor(experiment, levels = experiment_order),
    Cell.type = factor(Cell.type, levels = c("dI4_dILA", "dI5_dILB"))  # order: dI4 below
  ) %>%
  arrange(experiment, Cell.type)

# Step 3: Compute ymin/ymax manually for stacking
summary_ratios <- summary_ratios %>%
  group_by(experiment) %>%
  mutate(
    ymin = c(0, median_ratio[1]),
    ymax = ymin + median_ratio
  ) %>%
  ungroup()

# Step 4: Plot
ggplot(summary_ratios, aes(x = experiment, fill = Cell.type)) +
  # Manual stacked bars
  geom_rect(aes(xmin = as.numeric(experiment) - 0.3,
                xmax = as.numeric(experiment) + 0.3,
                ymin = ymin,
                ymax = ymax),
            color = "black") +

  # Error bars
  geom_errorbar(aes(
    x = experiment,
    ymin = pmax(ymax - sem, ymin),  # Prevents going below bottom of bar
    ymax = pmin(ymax + sem, 1)      # Prevents going over 1
  ),
  width = 0.2,
  color = "black",
  position = position_identity()) +

  # Ratio labels
  geom_text(aes(
    x = experiment,
    y = ymin + median_ratio / 2,
    label = round(median_ratio, 2)
  ),
  color = "white",
  size = 4) +

  # Manual fill colors
  scale_fill_manual(
    values = c("dI4_dILA" = "#3F7D83", "dI5_dILB" = "#BA600F"),
    breaks = c("dI4_dILA", "dI5_dILB")
  ) +

  # Y axis config
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    expand = c(0, 0)
  ) +

  # X axis config
  scale_x_discrete() +

  # Labels and theme
  labs(
    title = "Stacked median ratios of dI4 and dI5 in all cells",
    x = NULL,
    y = "Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold", margin = margin(b = 20)),
    panel.grid.minor.y = element_blank(),
    legend.title = element_blank(),
    plot.margin = margin(t = 30, r = 15, b = 15, l = 15)
  )


ggsave(filename = "dI4_dI5_ratio_all_median_replicates.pdf", plot = last_plot(), path = plots,
    scale = 1, width = 15 ,
    height = 16, units = "cm",
    dpi = 300, bg = "transparent")

```

### Distribution of dI4/dI5 in clones
```{r dI4-dI5-KDE}
# Plots KDE and histogram of dI4 fraction within clones containing both dI4 and dI5 cells

library(dplyr)
library(ggplot2)
library(tidyr)

# Step 1: Filter dI4/dI5 cells from multi-cell clones only
di_cells <- seur_obj@meta.data %>%
  filter(
    Cell.type %in% c("dI4_dILA", "dI5_dILB"),
    !is.na(exp_clone),
    !is.na(experiment),
    multi_cell_clone == "multi-cell_clone"  # NEW filter here
  ) %>%
  select(exp_clone, Cell.type, experiment)

# Step 2: Count dI4/dI5 cells per clone per experiment
clone_counts <- di_cells %>%
  group_by(exp_clone, experiment, Cell.type) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = Cell.type, values_from = n, values_fill = 0) %>%
  filter(`dI4_dILA` + `dI5_dILB` > 0) %>%
  mutate(ratio = `dI4_dILA` / (`dI4_dILA` + `dI5_dILB`))


# Step 3: Calculate median ratio per experiment
medians <- clone_counts %>%
  group_by(experiment) %>%
  summarise(med = median(ratio), .groups = "drop")

# Step 2: #KDE and histogram
ggplot(clone_counts, aes(x = ratio)) +
  geom_histogram(aes(y = ..density..),
                 binwidth = 0.05,
                 fill = "lightgray",
                 color = "black",
                 alpha = 0.5) +
  geom_density(fill = "cornflowerblue", alpha = 0.4, color = NA) +
  # geom_vline(data = medians, aes(xintercept = med),
  #            color = "red", linetype = "dashed", linewidth = 1) +
  # facet_wrap(~ experiment, nrow = 1) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Fraction dI4 / (dI4 + dI5) in clones",
    y = "Density"
  )

ggsave(filename = "dI4_dI5_KDE_tot.pdf", plot = last_plot(), path = plots,
    scale = 1, width = 18,
    height = 15, units = "cm",
    dpi = 300, bg = "transparent")


ggplot(clone_counts, aes(x = ratio)) +
  geom_histogram(binwidth = 0.1,
                 fill = "lightgray",
                 color = "black") +
  theme_minimal(base_size = 14) +
  labs(
    x = "Fraction dI4 / (dI4 + dI5) in clones",
    y = "Count"
  ) +
  facet_wrap(~ experiment, nrow = 1)
```


## S9
### UMAP by experiment S9
```{r umap-by-experiment}
# Plots UMAPs coloured by experiment (combined and split) for Fig. S7

colors <- c("violet", "steelblue4")

seur_obj$experiment <- factor(seur_obj$experiment, levels = c("CS16", "CS17"))

DimPlot(seur_obj,
            label = F,
        group.by = "experiment",
        shuffle = T,
        pt.size= 1,
        dims = c(1, 3),
        cols = colors,
        repel = T
        ) +
      labs(title = NULL) +  # Removes title
      coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL) +
      guides(color = guide_legend(override.aes = list(size = 4))) + # Adjust legend appearance if needed
      theme(
        legend.position = "right",  # Forces legend to stay in a fixed position
        legend.key.size = unit(0.5, "cm"),  # Standardize legend key size
        legend.text = element_text(size = 12)  # Standardize legend text size
      ) +
  NoLegend()

ggsave(filename = "UMAP_samples_colors-nolegend.png", plot = last_plot(), path = plots,
    scale = 1, width = 23 ,
    height = 23, units = "cm",
    dpi = 300, bg = "transparent")

#UMAP by experiment

DimPlot(seur_obj,
            label = F,
        group.by = "experiment",
        split.by = "experiment",
        shuffle = T,
        pt.size= 1,
        dims = c(1, 3),
        cols = colors,
        repel = T
        ) +
      labs(title = NULL) +  # Removes title
      coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL) +
      guides(color = guide_legend(override.aes = list(size = 4))) + # Adjust legend appearance if needed
      theme(
        legend.position = "right",  # Forces legend to stay in a fixed position
        legend.key.size = unit(0.5, "cm"),  # Standardize legend key size
        legend.text = element_text(size = 12)  # Standardize legend text size
      )

ggsave(filename = "UMAP_samples_colors_split.png", plot = last_plot(), path = plots,
    scale = 1, width = 37 ,
    height = 23, units = "cm",
    dpi = 300, bg = "transparent")

```
### FeaturePlots
```{r featureplots-markers}
# Generates and saves FeaturePlots for each marker gene in a loop, skipping missing genes

##Featureplot multiple features in loop


markers = c("OLIG2", "SOX2", "NKX6-1", "PAX6", "PAX7", "PDGFRA", "GFAP", "FOXJ1", "SLC17A6", "SLC32A1", "GAD1",  "SLC6A5")
markers = c("PAX2", "GATA3", "GATA2", "EN1", "EVX1", "VSX2", "SIM1")
markers = c("NKX6-2", "LHX2", "ISL1", "FOXD3", "OLIG1")


#If you have the gene names directly:

for (gene in markers) {
  # Try to generate and save the plot
  tryCatch({
    # Generate the plot
    p <- FeaturePlot(seur_obj, features = gene,
                     dims = c(1, 3),
                     min.cutoff = "q5", max.cutoff = "q90",
                     cols = c("grey80", "dodgerblue4"),
                     order = TRUE) +
         scale_x_discrete(breaks = NULL) +
         scale_y_discrete(breaks = NULL) +
         labs(title = paste0(gene, " expressing cells"))

    print(p)  # Ensure the plot is evaluated

    # Save the plot
    ggsave(
      filename = paste0(gene, "_q5q90Featureplot-dims1-2.png"),
      plot = p,
      path = plots,
      scale = 1,
      width = 17,
      height = 16,
      units = "cm",
      dpi = 300,
      bg = "transparent"
    )
  },
  error = function(e) {
    message(paste("Skipping", gene, "– not found or plot failed."))
  })
}


```

## S10
### Plot clones detected and multi-cell clones over UMAP S8
```{r umap-detection-multicell}
# Plots UMAPs showing LARRY barcode detection status and multi-cell clone classification for Fig. S8

Meta_Highlight_Plot(seurat_object = seur_obj,pt.size = 1.2, dims = c(1, 3), meta_data_column = "clone_id_detected",
    meta_data_highlight = c("not_detected","detected"),
    highlight_color = c(  "darkcyan","#E8997A", "grey90"),
    background_color = "grey90")+
        coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL)+
      guides(color = guide_legend(override.aes = list(size = 4))) + # Adjust legend appearance if needed
      theme(
        legend.position = "right",  # Forces legend to stay in a fixed position
        legend.key.size = unit(0.5, "cm"),  # Standardize legend key size
        legend.text = element_text(size = 12)  # Standardize legend text size
      ) +
  NoLegend()
ggsave(filename = "clone_detected-noledeng.png", plot = last_plot(), path = plots,
    scale = 1, width = 23 ,
    height = 23, units = "cm",
    dpi = 300, bg = "transparent")




DimPlot(seur_obj,
        pt.size= 1,
        cols = c("multi-cell_clone" = "darkblue", "single-cell_clone" = "red"),
        shuffle = T,
        dims = c(1,3),
        order = T,
        na.value = "grey90",
        group.by = "multi_cell_clone") +
      labs(title = NULL) +  # Removes title
      coord_fixed(ratio = 1) +
      scale_x_discrete(breaks = NULL) +
      scale_y_discrete(breaks = NULL) +
      guides(color = guide_legend(override.aes = list(size = 4))) + # Adjust legend appearance if needed
      theme(
        legend.position = "right",  # Forces legend to stay in a fixed position
        legend.key.size = unit(0.5, "cm"),  # Standardize legend key size
        legend.text = element_text(size = 12)  # Standardize legend text size
      )+
  NoLegend()


ggsave(filename = "multicellular_clones-nolegend.png", plot = last_plot(), path = plots,
    scale = 1, width = 23 ,
    height = 23, units = "cm",
    dpi = 300, bg = "transparent")

```


### Pie chart — number of cells in multi-cell clones S8
```{r pie-cells-in-multicell-clones}
# Plots a pie chart of the proportion of cells belonging to multi-cell vs single-cell clones

# Step 1: Prepare data (removing NA)
clone_counts <- seur_obj@meta.data %>%
  filter(!is.na(multi_cell_clone)) %>%
  count(multi_cell_clone) %>%
  mutate(
    per = round(n / sum(n) * 100, 1),
    label = paste0(per, "%")
  )

# Step 2: Plot
ggplot(data = clone_counts) +
  geom_bar(aes(x = "", y = per, fill = multi_cell_clone), stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  geom_text(aes(x = 1, y = cumsum(per) - per / 2, label = label), color = "white", size = 6) +
  scale_fill_manual(
    labels = c("Multicellular clones", "Single-cell clones"),
    name = NULL,
    values = c("multi-cell_clone" = "blue", "single-cell_clone" = "red")
  )

# Step 3: Save plot
ggsave(
  filename = "multicell_clones_cells_pie.pdf",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 26,
  height = 15,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)
```

### Pie chart — number of multi-cell clones S8
```{r pie-count-multicell-clones}
# Plots a pie chart of the number of distinct multi-cell vs single-cell clones

# Step 1: Filter out NA exp_clone values and group by multi_cell_clone
clone_summary <- seur_obj@meta.data %>%
  filter(!is.na(multi_cell_clone), !is.na(exp_clone)) %>%
  group_by(multi_cell_clone) %>%
  summarise(n_clones = n_distinct(exp_clone)) %>%
  ungroup()

print(clone_summary)


# Add percentages and labels
clone_summary <- clone_summary %>%
  mutate(
    per = round(n_clones / sum(n_clones) * 100, 1),
    label = paste0(per, "%")
  )

# Plot
ggplot(data = clone_summary) +
  geom_bar(aes(x = "", y = per, fill = multi_cell_clone), stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  geom_text(aes(x = 1, y = cumsum(per) - per / 2, label = label), color = "white", size = 6) +
  scale_fill_manual(
  name = NULL,
  values = c("multi-cell_clone" = "blue", "single-cell_clone" = "red"),
  labels = c("multi-cell_clone" = "Multicellular clones",
             "single-cell_clone" = "Single-cell clones")
  )


# Save
ggsave(
  filename = "multicell_vs_single_clones_count_pie.pdf",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 26,
  height = 15,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)
```


### Pie charts — clone detected S8
```{r pie-clone-detection}
# Plots pie charts of LARRY detection rates overall and faceted by experiment for Fig. S8

clone_detect_counts <- seur_obj@meta.data %>%
  filter(!grepl("control", samples_LARRY, ignore.case = TRUE)) %>%  # remove control samples
  filter(!is.na(clone_id_detected)) %>%  # remove NAs
  count(clone_id_detected) %>%
  mutate(
    per = round(n / sum(n) * 100, 1),
    label = paste0(per, "%")
  )

ggplot(clone_detect_counts) +
  geom_bar(aes(x = "", y = per, fill = clone_id_detected), stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  geom_text(
    aes(x = 1, y = cumsum(per) - per / 2, label = label),
    color = "white",
    size = 6
  ) +
  scale_fill_manual(
    labels = c("Detected", "Not detected"),
    name = NULL,
    values = c("detected" = "darkcyan", "not_detected" = "#E8997A")
  )

ggsave(
  filename = "clone_detected_avg_pie.pdf",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 18,
  height = 15,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)


# Step 1: Prepare counts per experiment, excluding controls
clone_detect_facet <- seur_obj@meta.data %>%
  filter(!grepl("control", samples_LARRY, ignore.case = TRUE)) %>%
  filter(!is.na(clone_id_detected), !is.na(experiment)) %>%
  count(experiment, clone_id_detected) %>%
  group_by(experiment) %>%
  mutate(
    per = round(n / sum(n) * 100, 1),
    label = paste0(per, "%"),
    ypos = cumsum(per) - per / 2
  ) %>%
  ungroup()

ggplot(clone_detect_facet, aes(x = "", y = per, fill = clone_id_detected)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  geom_text(aes(y = ypos, label = label), color = "white", size = 4) +
  coord_polar("y", start = 0) +
  scale_fill_manual(
    values = c("detected" = "darkcyan", "not_detected" = "#E8997A"),
    labels = c("Detected", "Not detected"),
    name = NULL
  ) +
  facet_wrap(~ experiment) +
  theme_void() +
  theme(
    strip.text = element_text(size = 12, face = "bold")
  )

# Step 3: Save plot
ggsave(
  filename = "clone_detected_experiments_pie.pdf",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 45,
  height = 15,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)


```


### UpSet plot — broad clone categories
```{r upset-broad-categories}
# Plots an UpSet plot of neuron vs neural progenitor clonal co-occurrence in multi-cell clones

seur_obj@meta.data <- seur_obj@meta.data %>%
  mutate(cell_category = case_when(
    Cell.type %in% c("RP_dp1", "dp1-3", "dpL", "dp6/p2", "pMN/p3", "FP") ~ "Neural Progenitors",
    Cell.type %in% c("dI1", "dI2/3", "dI4_dILA", "dI5_dILB", "dI6", "V0", "V1", "V2",  "MN", "V3") ~ "Neurons"
  ))

#
# celltype_order <- c("RP-dp1","dp1-3", "dpL", "dp6/p2", "pMN/p3",  "FP", "OPCs", "dI1" , "dI2/3","dI4_dILA","dI5_dILB", "dI6","V0","V1","V2","MN","V3", "Unknown", "mesoderm/drg")

table(seur_obj@meta.data$cell_category)



library(ComplexUpset)

upset_df <- seur_obj@meta.data %>%
  filter(
    !is.na(exp_clone),
    !is.na(cell_category),
    multi_cell_clone == "multi-cell_clone"
  ) %>%
  distinct(exp_clone, cell_category) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = cell_category,
    values_from = present,
    values_fill = list(present = 0)
  )

upset_df

# Plot with custom color palette and clean theme

upset(
  upset_df,
  intersect = c("Neurons", "Neural Progenitors"),  # Custom order
  name = "Clonal Sharing",
  set_sizes = FALSE,
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
)


ggsave(filename = "Upset_plots_broad_cat.pdf", plot = last_plot(), path = plots,
    scale = 1, width = 15 ,
    height = 12, units = "cm",
    dpi = 300, bg = "transparent")



```

### Count dorsal and ventral clones (text)
```{r count-dorsal-ventral-clones}
# Counts the number of single-cell clones that are exclusively dorsal or exclusively ventral

# Define classes
# --- Define dorsal/ventral ---
dorsal <- c("RP-dp1", "dp1-3", "dpL", "dI1", "dI2/3", "dI4_dILA", "dI5_dILB")
ventral <- c("dp6/p2", "pMN/p3", "FP", "OPCs", "dI6", "V0", "V1", "V2", "MN", "V3")

# --- Filter metadata and assign class ---
meta_classed <- seur_obj@meta.data %>%
  rownames_to_column("cell") %>%
  filter(!is.na(exp_clone), multi_cell_clone == "single-cell_clone") %>%
  mutate(class = case_when(
    Cell.type %in% dorsal ~ "dorsal",
    Cell.type %in% ventral ~ "ventral",
    TRUE ~ "other"
  ))

# Re-classify clones while ignoring "other"
pure_clones <- meta_classed %>%
  filter(class %in% c("dorsal", "ventral")) %>%  # exclude "other" from this check
  group_by(exp_clone) %>%
  summarise(class_set = list(unique(class))) %>%
  filter(lengths(class_set) == 1) %>%
  mutate(class = unlist(class_set)) %>%
  select(exp_clone, class)

# Count pure clones
clone_counts <- pure_clones %>%
  count(class, name = "n_clones")

clone_counts

```


### Clonal size
```{r clone-size-violin}
# Plots violin plots of clone size for all clones and multicell-only clones, stratified by experiment

# Step 1: Prepare clone sizes with experiment info
clone_sizes <- seur_obj@meta.data %>%
  filter(!is.na(exp_clone), !is.na(experiment)) %>%
  group_by(exp_clone) %>%
  summarise(
    size = n(),
    experiment = first(experiment)
  )

# Step 2: Set experiment order manually (HH28, HH31, HH35)
clone_sizes$experiment <- factor(clone_sizes$experiment, levels = c("CS16", "CS17"))

# Step 3: Plot violin plot with custom colors, no points, and median value
colors <- c("violet", "steelblue4")

ggplot(clone_sizes, aes(x = experiment, y = size, fill = experiment)) +
  geom_violin(trim = TRUE, scale = "width", color = NA) +
  stat_summary(fun = mean, geom = "point", shape = 95, size = 6, color = "black") +
  stat_summary(fun = mean, geom = "text", aes(label = round(..y.., 2)),
               vjust = -1.2, size = 5, color = "black") +
  scale_fill_manual(values = colors) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Clone Sizes Across Experiments",
    x = "Experiment",
    y = "Clone Size (number of cells)"
  ) +
  theme(legend.position = "none")

ggsave(filename = "n_cells_all_clones_vln.pdf", plot = last_plot(), path = plots,
    scale = 1, width = 20 ,
    height = 17, units = "cm",
    dpi = 300, bg = "transparent")


---

  # Step 1: Filter to multicellular clones and calculate clone sizes
clone_sizes_multi <- seur_obj@meta.data %>%
  filter(
    !is.na(exp_clone),
    !is.na(experiment),
    multi_cell_clone == "multi-cell_clone"
  ) %>%
  group_by(exp_clone) %>%
  summarise(
    size = n(),
    experiment = first(experiment)
  )

# Step 2: Set experiment order manually
clone_sizes$experiment <- factor(clone_sizes$experiment, levels = c("CS16", "CS17"))

# Step 3: Plot violin plot with custom colors, no points, and mean value
colors <- c("violet", "steelblue4")

ggplot(clone_sizes_multi, aes(x = experiment, y = size, fill = experiment)) +
  geom_violin(trim = TRUE, scale = "width", color = NA) +
  stat_summary(fun = mean, geom = "point", shape = 95, size = 6, color = "black") +
  stat_summary(fun = mean, geom = "text", aes(label = round(..y.., 2)),
               vjust = -1.2, size = 5, color = "black") +
  scale_fill_manual(values = colors) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Multicellular Clone Sizes Across Experiments",
    x = "Experiment",
    y = "Clone Size (number of cells)"
  ) +
  theme(legend.position = "none")+
 coord_cartesian(ylim = c(1, NA))



ggsave(filename = "n_cells_multicell_clones_vln.pdf", plot = last_plot(), path = plots,
    scale = 1, width = 20 ,
    height = 17, units = "cm",
    dpi = 300, bg = "transparent")


```


### Plot biggest shared clones over UMAP
```{r umap-top10-shared-clones}
# Highlights the top 10 largest clones containing both dI4 and dI5 cells on the UMAP

# 1. Get clone counts
clone_counts <- di_cells %>%
  group_by(exp_clone) %>%
  summarise(
    dI4 = sum(Cell.type == "dI4_dILA"),
    dI5 = sum(Cell.type == "dI5_dILB"),
    total = n(),
    .groups = "drop"
  )

# 2. Filter for shared clones (both dI4 + dI5) and select top 10 by size
top_shared <- clone_counts %>%
  filter(dI4 > 0 & dI5 > 0) %>%
  arrange(desc(total)) %>%
  slice_head(n = 10) %>%
  pull(exp_clone)

# 3. Get UMAP coordinates
umap_coords <- Embeddings(seur_obj, "umap")[, c(1,3)] %>% as.data.frame()
umap_coords$cell <- rownames(umap_coords)

meta_df <- seur_obj@meta.data %>%
  tibble::rownames_to_column("cell")

plot_df <- left_join(umap_coords, meta_df, by = "cell")

# 4. Define highlight column
plot_df <- plot_df %>%
  mutate(highlight = ifelse(exp_clone %in% top_shared, exp_clone, "Other"))

# 5. Plot: grey background + colored highlights
ggplot() +
  geom_point(
    data = plot_df %>% filter(highlight == "Other"),
    aes(x = UMAP_1, y = UMAP_3),
    color = "grey90", size = 0.5, alpha = 0.5
  ) +
  geom_point(
    data = plot_df %>% filter(highlight != "Other"),
    aes(x = UMAP_1, y = UMAP_3, color = highlight),
    size = 3, alpha = 0.9
  ) +
  scale_color_manual(
    values = setNames(RColorBrewer::brewer.pal(10, "Paired"), top_shared)
  ) +
  labs(title = "Top 10 largest shared dI4/dI5 clones on UMAP") +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggsave(
  filename = "top10_clones_dI4-dI5shared.png",
  plot = last_plot(),
  path = plots,
  scale = 1,
  width = 20,
  height = 16,
  units = "cm",
  dpi = 300,
  bg = "transparent"
)



```

